---
title: "Untitled"
output: html_document
date: "2025-09-24"
---

# Covariate Selection and Censoring Adjustment for Under‑Five Mortality

```{r}
# This script does:
# 1. Loads 21 raw covariate rasters (from local paths – user must change)
# 2. Projects/resamples to 5 km grid
# 3. Removes 5 covariates with pairwise correlation >= 0.7 → keeps 16
# 4. Extracts covariate values at DHS cluster locations
# 5. Fits Random Survival Forest (RSF) with birth cohort stratification
# 6. Adjusts for informative censoring 
# 7. Saves censoring-adjusted mortality variable

# IMPORTANT – DATA NOT INCLUDED
# This script requires restricted DHS data for Ethiopia.
# The data are NOT included in this deposit.
# To run the script, you must:
#   1. Obtain the DHS file 'all apend.dta' from https://dhsprogram.com
#   2. Change the paths in the 'step1' chunk to point to your local files.
# The script is provided for methodological transparency.
```

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Core data processing
# Core data processing
library(haven)        # Read DTA files
library(terra)        # Spatial raster processing
library(sf)           # Spatial vector operations
library(dplyr)        # Data manipulation
library(tidyr)        # Data reshaping
library(purrr)        # Functional programming
library(moments)      # Skewness and kurtosis
library(corrr)        # Correlation matrices
library(jsonlite)     # JSON metadata export
library(writexl)      # Write Excel files
library(exactextractr) # Extract raster values over polygons
# Survival analysis
library(survival)         # Survival objects and Cox models
library(randomForestSRC)  # Random Survival Forest (primary method)
library(pec)              # Prediction error curves
library(Hmisc)            # Misc functions including rcorr.cens
library(gbm)              # Gradient boosting survival model
# Visualization and diagnostics
library(ggplot2)      # Plotting
library(corrplot)     # Correlation plots
library(pROC)         # ROC analysis
library(timeROC)      # Time-dependent ROC
# Spatial and Bayesian modelling
library(INLA)         # Integrated nested Laplace approximations
library(fmesher)      # Mesh construction for INLA
library(e1071)        # Skewness calculation
# Multicollinearity diagnostics
library(car)          # Variance Inflation Factor (VIF)     
# Functional programming
# Set seed for reproducibility
set.seed(123)
```


## 1.  Data Preparation, Covariate Processing, and Censoring Adjustment


```{r step1}
# IMPORTANT: The following paths are for the author's local machine.
# You MUST change them to point to your own DHS data file,
# covariate rasters, and output directories before running.
DHS_DATA_PATH <- "D:/new resea/DHS/Ethiopia/all apend.dta"
COVARIATE_PATH <- "D:/Ethi covaraite raster/Ethiopia/"
OUTPUT_BASE <- "D:/new resea/Ethiopia/"
# Read DHS data using the variable
data <- read_dta(DHS_DATA_PATH)
# Define output directories
CENSOR_DIR <- paste0(OUTPUT_BASE, "Mean for corrlation and censoring/")
CENSOR_CENSOR_DIR <- paste0(CENSOR_DIR, "censor/")
COVARIATE_STACK_DIR <- paste0(OUTPUT_BASE, "covariate_stack/")
# Create directories if they don't exist
for(dir in c(CENSOR_DIR, CENSOR_CENSOR_DIR, COVARIATE_STACK_DIR)) {
  if(!dir.exists(dir)) dir.create(dir, recursive = TRUE)
}
```

### 1.1.  DHS Data Preparation and Outcome Construction

```{r step2-handle-missing-child-age}
# Replace missing child age using age at death variables.
data$b19[is.na(data$b19)] <- data$b8[is.na(data$b19)] * 12
data$b19[is.na(data$b19)] <- data$b7[is.na(data$b19)]
```

```{r step3-convert-ethiopian-calendar}
# Convert Ethiopian calendar birth year to Gregorian calendar.
eth_to_gc_year <- function(b2, b1) {
  gc_year <- ifelse(b1 >= 1 & b1 <= 4,
                    b2 + 8,
                    b2 + 7)
  return(gc_year)
}
# Apply to DHS dataframe
data$gc_birth_year <- eth_to_gc_year(data$b2, data$b1)
```

```{r step4-create-cluster-id}
# Create unique cluster identifier combining birth year, survey year, and cluster number.
data$Clu_id <- paste0(data$gc_birth_year, data$year, data$v001)
table(data$b5)
```

```{r step5-recode-mortality-outcome}
# Recode mortality variable so that death = 1 and survival = 0.

table(data$b5)
data$b5 <- ifelse(data$b5 == 0, 1,
                  ifelse(data$b5 == 1, 0, data$b5))
table(data$b5)
```

```{r step6-aggregate-cluster-counts}
#  Calculate sample size and deaths per cluster and restrict births to year ≥ 2000.
# number of sample(n) in each id
data$b <- 1
data <- data %>%
  group_by(Clu_id) %>%
  mutate(n = sum(b)) %>%
  ungroup()
# number of cases (y) in each id
data <- data %>%
  group_by(Clu_id) %>%
  mutate(y = sum(b5)) %>%
  ungroup()
data <- data[!duplicated(data$Clu_id), ]
data <- data[data$gc_birth_year >= 2000, ]
table(data$gc_birth_year)
```


### 1.2. Spatial Covariate Processing and Grid Harmonization

```{r step7-load-raster-covariates}
# Load environmental, climatic, and socio-demographic raster covariates.

CHI   <- rast(paste0(COVARIATE_PATH, "chrips/mean_1.tif"))
DEAR  <- rast(paste0(COVARIATE_PATH, "deareha/dear_mea.tif"))
DPT   <- rast(paste0(COVARIATE_PATH, "dpt3/DPT_mean.tif"))
EBF   <- rast(paste0(COVARIATE_PATH, "EBF/mean_3.tif"))
ELEV  <- rast(paste0(COVARIATE_PATH, "elevation/ELEVATION.tif"))
ERA5  <- rast(paste0(COVARIATE_PATH, "ERA5 Ethiopia/ERA_mean.tif"))
HIV   <- rast(paste0(COVARIATE_PATH, "HIV/HIV_mean.tif"))
LRI   <- rast(paste0(COVARIATE_PATH, "LRI/LDI_mean.tif"))
MALA  <- rast(paste0(COVARIATE_PATH, "malaria/malare_mean.tif"))
NDVI  <- rast(paste0(COVARIATE_PATH, "NDVI Ethiopia/NDVI_mean1.tif"))
NIGH  <- rast(paste0(COVARIATE_PATH, "nightlight/mean_8.tif"))
PM25  <- rast(paste0(COVARIATE_PATH, "pm2.5/PM2.5._mean.tif"))
SW    <- rast(paste0(COVARIATE_PATH, "Sever wasting/mean_11.tif"))
SLO   <- rast(paste0(COVARIATE_PATH, "slop/eth_slope_merit103_100m_v1.tif"))
STUN  <- rast(paste0(COVARIATE_PATH, "stunting/mean_12.tif"))
WASH  <- rast(paste0(COVARIATE_PATH, "wash/mean_13.tif"))
WATE  <- rast(paste0(COVARIATE_PATH, "water/mean_14.tif"))
EDU1  <- rast(paste0(COVARIATE_PATH, "education.tif"))
GRID  <- rast(paste0(COVARIATE_PATH, "grdi.tif"))
ROAD  <- rast(paste0(COVARIATE_PATH, "road density Ethiopia.tif"))
HF    <- rast(paste0(COVARIATE_PATH, "travle time to hf.tif"))

cat("All covariate rasters loaded successfully\n")
```

```{r step8-inspect-education-raster}
# Inspect education raster layers and compute mean excluding incomplete years.

nlyr(EDU1)
plot(EDU1$education_1)
plot(EDU1$education_2)
plot(EDU1$education_3)
plot(EDU1$education_4)
plot(EDU1$education_5)
plot(EDU1$education_6)
plot(EDU1$education_7)
plot(EDU1$education_8)
plot(EDU1$education_9)
plot(EDU1$education_10)
plot(EDU1$education_11)
plot(EDU1$education_12)
plot(EDU1$education_13)
plot(EDU1$education_14)
plot(EDU1$education_15)
plot(EDU1$education_16)
plot(EDU1$education_17)
plot(EDU1$education_18)
plot(EDU1$education_19)
EDU <- mean(EDU1[[1:18]])   # use only the first 18 layers  
plot(EDU)
```

```{r step9-project-rasters-to-utm}
# Project all rasters to UTM Zone 37N coordinate system.
target_crs <- "EPSG:32637"  # UTM zone Ethiopia
CHI_utm  <- project(CHI, target_crs)
DEAR_utm  <- project(DEAR, target_crs)
DPT_utm  <- project(DPT, target_crs)
EBF_utm  <- project(EBF, target_crs)
ELEV_utm <- project(ELEV, target_crs)
ERA5_utm <- project(ERA5, target_crs)
HIV_utm  <- project(HIV,  target_crs)
LRI_utm <- project(LRI, target_crs)
MALA_utm <- project(MALA, target_crs)
NDVI_utm <- project(NDVI, target_crs)
NIGH_utm   <- project(NIGH,   target_crs)
PM25_utm  <- project(PM25,  target_crs)
SW_utm <- project(SW, target_crs)
SLO_utm <- project(SLO, target_crs)
STUN_utm  <- project(STUN,  target_crs)
WASH_utm <- project(WASH, target_crs)
WATE_utm <- project(WATE, target_crs)
EDU_utm <- project(EDU, target_crs)
GRID_utm <- project(GRID, target_crs)
ROAD_utm <- project(ROAD, target_crs)
HF_utm <- project(HF, target_crs)
```

```{r step10-create-5km-template}
# Create a template raster with 5×5 km spatial resolution.
template_5km <- rast(ext(CHI_utm), res = 5000, crs = target_crs)
```

```{r step11-resample-covariates}
#  Resample all covariates to the 5 km template grid.

Chi5km   <- resample(CHI_utm,   template_5km, method = "bilinear")
dear5km  <- resample(DEAR_utm,  template_5km, method = "bilinear")
DPT5km  <- resample(DPT_utm,  template_5km, method = "bilinear")
Eluv5km  <- resample(ELEV_utm,  template_5km, method = "bilinear")
Slo5km   <- resample(SLO_utm,   template_5km, method = "bilinear")
era55km  <- resample(ERA5_utm,  template_5km, method = "bilinear")
Hiv5km  <- resample(HIV_utm,  template_5km, method = "bilinear")
Lri5km  <- resample(LRI_utm,  template_5km, method = "bilinear")
Plas5km  <- resample(MALA_utm,  template_5km, method = "bilinear")
Ndvi5km  <- resample(NDVI_utm,  template_5km, method = "bilinear")
Pm255km  <- resample(PM25_utm,  template_5km, method = "bilinear")
SW5km    <- resample(SW_utm,    template_5km, method = "bilinear")
Stu5km   <- resample(STUN_utm,   template_5km, method = "bilinear")
Wash5km  <- resample(WASH_utm,  template_5km, method = "bilinear")
Wate5km  <- resample(WATE_utm,  template_5km, method = "bilinear")
ebf5km  <- resample(EBF_utm,  template_5km, method = "bilinear")
nili5km  <- resample(NIGH_utm,  template_5km, method = "bilinear")
Edu5km   <- resample(EDU_utm,   template_5km, method = "bilinear")
Gdid5km  <- resample(GRID_utm,  template_5km, method = "bilinear")
Road5km  <- resample(ROAD_utm,  template_5km, method = "bilinear")
Hf5km  <- resample(HF_utm,  template_5km, method = "bilinear")
```

```{r step12-check-raster-consistency}
# Verify resolution, number of cells, layers, and visualize rasters.
res(ebf5km)
res(nili5km)
res(Chi5km)  # should be 5000 x 5000 meters
res(dear5km)  # should be 5000 x 5000 meters
res(Eluv5km)  # should be 5000 x 5000 meters
res(Slo5km)  # should be 5000 x 5000 meters
res(era55km)  # should be 5000 x 5000 meters
res(Plas5km)  # should be 5000 x 5000 meters
res(Pm255km)  # should be 5000 x 5000 meters
res(SW5km)  # should be 5000 x 5000 meters
res(Stu5km)  # should be 5000 x 5000 meters
res(Wash5km)  # should be 5000 x 5000 meters
res(Wate5km)  # should be 5000 x 5000 meters
res(Hf5km)
res(Road5km)
res(Gdid5km)
res(Edu5km)
res(DPT5km)
res(Hiv5km)
res(Lri5km)
res(Ndvi5km)
ncell(ebf5km)
ncell(nili5km)
ncell(Wate5km) 
ncell(Wash5km) 
ncell(Stu5km)
ncell(SW5km) 
ncell(Pm255km) 
ncell(Plas5km) 
ncell(era55km) 
ncell(Slo5km) 
ncell(Eluv5km) 
ncell(dear5km) 
ncell(Chi5km) 
ncell(Hf5km)
ncell(Road5km)
ncell(Gdid5km)
ncell(Edu5km)
ncell(DPT5km)
ncell(Hiv5km)
ncell(Lri5km)
ncell(Ndvi5km)
nlyr(ebf5km); nlyr(nili5km); nlyr(Chi5km); nlyr(dear5km); 
nlyr(Eluv5km); nlyr(Slo5km); nlyr(era55km); 
nlyr(Plas5km); nlyr(Pm255km); nlyr(SW5km); 
nlyr(Stu5km); nlyr(Wash5km); nlyr(Wate5km); nlyr(Hf5km); nlyr(Road5km); nlyr(Gdid5km); nlyr(Edu5km); nlyr(DPT5km); nlyr(Hiv5km); nlyr(Lri5km); nlyr(Ndvi5km)
plot(ebf5km[[1]])
plot(nili5km[[1]])
plot(Wate5km[[1]]) 
plot(Wash5km[[1]]) 
plot(Stu5km[[1]])
plot(SW5km[[1]]) 
plot(Pm255km[[1]]) 
plot(Plas5km[[1]]) 
plot(era55km[[1]]) 
plot(Slo5km[[1]]) 
plot(Eluv5km[[1]]) 
plot(dear5km[[1]]) 
plot(Chi5km[[1]]) 
```

```{r step13-stack-and-export-covariates}
# Stack all raster covariates, assign names, and export them.
all_covariates_5km <- c(ebf5km[[1]], nili5km[[1]], Chi5km[[1]], dear5km[[1]], 
                        Eluv5km[[1]], Slo5km[[1]], era55km[[1]],
                        Plas5km[[1]], Pm255km[[1]], SW5km[[1]], Stu5km[[1]], 
                        Wash5km[[1]], Wate5km[[1]], Hf5km[[1]], Road5km[[1]], 
                        Gdid5km[[1]], Edu5km[[1]], DPT5km[[1]], Hiv5km[[1]], 
                        Lri5km[[1]], Ndvi5km[[1]])

names(all_covariates_5km) <- c("ebf", "nili", "Chi", "dear", "Eluv", "Slo", "era5",
                                "Plas", "Pm25", "SW", "Stu", "Wash", "Wate", 
                                "hf", "road", "grid", "educ", "dpt", "hiv", 
                                "lri", "ndvi")

# Set output folder
out_folder <- paste0(OUTPUT_BASE, "mean for prediction/mean5km_layers/")
if(!dir.exists(out_folder)) dir.create(out_folder, recursive = TRUE)

# Export each layer
for (i in 1:nlyr(all_covariates_5km)) {
  layer_name <- names(all_covariates_5km)[i]
  out_file <- paste0(out_folder, layer_name, ".tif")
  writeRaster(all_covariates_5km[[i]], filename = out_file, overwrite = TRUE)
}

# Save full stack
writeRaster(all_covariates_5km, 
            paste0(CENSOR_DIR, "all_covariates_5km.tif"), 
            overwrite = TRUE)

cat("  Covariate stack saved with", nlyr(all_covariates_5km), "layers\n")
cat("   Individual layers saved to:", out_folder, "\n")
```

###  1.3. Covariate Selection Using Correlation Analysis
Identify and remove highly correlated covariates before modeling.
All spatial covariates for Ethiopia, including environmental, health, and socio-demographic variables, were projected to UTM Zone 37N and resampled to a 5 × 5 km grid. Multi-year rasters, such as educational attainment, were summarized using pixel-wise means to represent long-term average conditions. This standardized grid ensured alignment across all covariates and facilitated variable selection via pairwise correlation and handling of censored outcome data. All raster processing and aggregation were performed independently by the author to ensure reproducibility and consistency.

```{r step14-compute-correlation-matrix}
# Extract raster values and compute correlation matrix.
names(all_covariates_5km)
cov_values <- values(all_covariates_5km)  # extract cell values
cov_values <- na.omit(cov_values)         # remove NAs
cor_matrix <- cor(cov_values, method = "pearson")  # or "spearman"
# Identify highly correlated pairs
high_corr <- findCorrelation(cor_matrix, cutoff = 0.7)  # adjust cutoff
high_corr
cor_matrix
```

```{r step15-remove-highly-correlated-covariates}
# Identify and remove highly correlated covariates.
cov_values_reduced <- cov_values[, -high_corr]
cor_matrix_reduced <- cor(cov_values_reduced)
cor_matrix_reduced

# Map short names from raster stack to readable labels
new_names <- c(
  "ebf" = "Exclusive breastfeeding",
  "Stu"="Stunting",
  "dpt"="DPT3 immunization",
  "ndvi"="NDVI",
  "nili" = "Night light intensity", 
  "Chi"   = "CHIRPS rainfall",
  "dear"  = "Diarrhea",
  "Eluv"  = "Elevation",
  "Slo"   = "Slope",
  "era5"  = "ERA5 climate",  
  "Plas"  = "Plasmodium falciparum",
  "Pm25"  = "PM2.5 air pollution",
  "SW"    = "Severe wasting",
  "Wash"  = "Sanitation facility",
  "Wate"  = "Water access",
   "hf"  = "Distance to health facility", 
  "road" =  "Road density",
  "grid" = "Grid index", 
  "educ" = "Educational attainment",
  "hiv"  = "HIV prevalence",
  "lri" = "Lower respiratory infection")
  old_names <- colnames(cor_matrix_reduced)
new_names_vec <- new_names[old_names]

colnames(cor_matrix_reduced) <- new_names_vec
rownames(cor_matrix_reduced) <- new_names_vec

cor_matrix_reduced

# all covaraites after removing the corlated variable (EBF, Stunting, DPT3, CHIRPS rainfall,)
covar5km_Cens_SHAP <- all_covariates_5km[[ -high_corr]]
# Check the layers
names(covar5km_Cens_SHAP)

writeRaster(covar5km_Cens_SHAP,
            paste0(CENSOR_DIR, "covar5km_Cens_SHAP.tif"),
            overwrite = TRUE)
res(all_covariates_5km)  # should be 5000 x 5000 meters
ncell(all_covariates_5km) 
ext(all_covariates_5km)
```


```{r step16-visualize-correlation}
# Save correlation plot to file instead of opening window
png(filename = paste0(CENSOR_DIR, "correlation_plot.png"),
    width = 10, height = 10, units = "in", res = 300)

corrplot(cor_matrix_reduced,
         method = "color",
         addCoef.col = "black",
         order = "hclust",
         type = "lower",
         tl.cex = 0.8,
         tl.srt = 45,
         tl.col = "black",
         cl.cex = 0.8,
         number.cex = 0.6)

dev.off()
cat("✅ Correlation plot saved to:", paste0(CENSOR_DIR, "correlation_plot.png"), "\n")
```

```{r step17-final-covariate-selection}
# Finalize the set of covariates used for modeling.
# Load the stack mean
covariates_5km <- rast(paste0(CENSOR_DIR, "all_covariates_5km.tif"))
# Layers to drop (spatial identifiers not used as covariates)
layers_to_drop <- c("grid", "road", "hf")
covariates_5km2 <- covariates_5km[[ !names(covariates_5km) %in% layers_to_drop ]]
# Extract values for correlation analysis
cov_val <- values(covariates_5km2)
cov_val <- na.omit(cov_val)

# Correlation matrix
cor_mat <- cor(cov_val, method = "pearson")
high_cor <- findCorrelation(cor_mat, cutoff = 0.7)

# Create final covariate stack
covar5km_fin <- covariates_5km2[[ -high_cor ]]
cat("✅ Final covariates selected:", paste(names(covar5km_fin), collapse=", "), "\n")

# Save final covariate stack
writeRaster(covar5km_fin,
            paste0(CENSOR_DIR, "covar5km_fin.tif"),
            overwrite = TRUE)
```


###  1.5. Construction of Survival Dataset for Censoring Adjustment

Prepare the child survival dataset and attach spatial covariates to each observation.


```{r step18-prepare-censoring-dataset}
# Prepare dataset used for survival-based censoring adjustment.
cova_5km_cens_SHAP <- rast(paste0(CENSOR_DIR, "covar5km_Cens_SHAP.tif"))
data <- read_dta(DHS_DATA_PATH)

# The died child missing their current age of child - replaced by age of death
data$b19[is.na(data$b19)] <- data$b8[is.na(data$b19)] * 12 
data$b19[is.na(data$b19)] <- data$b7[is.na(data$b19)]
eth_to_gc_year <- function(b2, b1) {
  gc_year <- ifelse(b1 >= 1 & b1 <= 4,
                    b2 + 8,
                    b2 + 7)
  return(gc_year)
}

# Apply to DHS dataframe
data$gc_birth_year <- eth_to_gc_year(data$b2, data$b1)
data$Clu_id <- paste0(data$gc_birth_year, data$year, data$v001)

# Recode mortality: 0 = alive, 1 = death
cat("\n📊 Original mortality distribution:\n")
table(data$b5)
data$b5 <- ifelse(data$b5 == 0, 1,
                  ifelse(data$b5 == 1, 0, data$b5))
cat("\n📊 Recoded mortality (1 = death):\n")
table(data$b5)

# Select only important variables
cens <- data[, c("b7", "b19", "b4", "v025", "v201", "b0", 
                 "Clu_id", "b5", "latnum", "longnum", "birth_year")]

cat("✅ Censoring dataset prepared with", nrow(cens), "rows\n")
```



```{r step19-extract-spatial-covariates-to-points}
# CONVERT DHS CLUSTERS TO SPATIAL POINTS AND EXTRACT RASTER VALUES
# (CORRECTED VERSION)

cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP: EXTRACTING RASTER COVARIATES FOR DHS CLUSTERS\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# 1. Load stacked raster
cat("\n📊 STEP 1: LOADING RASTER STACK\n")
cova_5km_cens_SHAP <- rast(paste0(CENSOR_DIR, "covar5km_Cens_SHAP.tif"))

# Check raster properties
cat("✅ Raster loaded successfully\n")
cat("  Number of layers:", nlyr(cova_5km_cens_SHAP), "\n")
cat("  Layer names:", paste(names(cova_5km_cens_SHAP), collapse=", "), "\n")
cat("  Raster CRS:", crs(cova_5km_cens_SHAP), "\n")

# Get extent
raster_ext <- ext(cova_5km_cens_SHAP)
cat("  Raster extent: xmin =", raster_ext[1], ", xmax =", raster_ext[2], 
    ", ymin =", raster_ext[3], ", ymax =", raster_ext[4], "\n")

# 2. Prepare DHS data
cat("\n📊 STEP 2: PREPARING DHS DATA\n")
cat("  Original data rows:", nrow(cens), "\n")

# Check for missing coordinates
missing_coords <- sum(is.na(cens$longnum) | is.na(cens$latnum))
if(missing_coords > 0) {
  cat("  ⚠️ Warning:", missing_coords, "rows have missing coordinates\n")
  cens_clean <- cens[!is.na(cens$longnum) & !is.na(cens$latnum), ]
  cat("  → Rows after removing missing coordinates:", nrow(cens_clean), "\n")
} else {
  cens_clean <- cens
  cat("  ✅ No missing coordinates\n")
}

# 3. Convert to spatial points
cat("\n📊 STEP 3: CREATING SPATIAL POINTS\n")
pts <- vect(cens_clean, geom = c("longnum", "latnum"), crs = "EPSG:4326")
cat("  ✅ Spatial points created:", nrow(pts), "\n")
cat("  Points CRS:", crs(pts), "\n")

# 4. Check CRS alignment
cat("\n📊 STEP 4: CHECKING CRS ALIGNMENT\n")
if(crs(pts) != crs(cova_5km_cens_SHAP)) {
  cat("  ⚠️ CRS mismatch - reprojecting points to match raster\n")
  pts <- project(pts, crs(cova_5km_cens_SHAP))
  cat("  ✅ Points reprojected. New CRS:", crs(pts), "\n")
} else {
  cat("  ✅ CRS match\n")
}

# 5. Extract values
cat("\n📊 STEP 5: EXTRACTING RASTER VALUES\n")
cat("  This may take a moment...\n")
vals <- terra::extract(cova_5km_cens_SHAP, pts)
cat("  ✅ Extraction complete\n")
cat("  Extracted values shape:", nrow(vals), "rows,", ncol(vals)-1, "columns\n")

# 6. Remove ID column
vals_data <- vals[, -1]

# 7. Check for NA values
na_count <- colSums(is.na(vals_data))
if(any(na_count > 0)) {
  cat("\n  ⚠️ Warning: Some extracted values are NA\n")
  na_summary <- data.frame(
    Layer = names(na_count),
    NA_count = na_count,
    NA_percent = round(100 * na_count / nrow(vals_data), 2)
  )
  print(na_summary[na_summary$NA_count > 0, ])
} else {
  cat("\n  ✅ No NA values in extracted data\n")
}

# 8. Combine with original data
cat("\n📊 STEP 6: COMBINING WITH ORIGINAL DATA\n")
cens_covariates <- cbind(cens_clean, vals_data)
cat("  ✅ Final dataset dimensions:", nrow(cens_covariates), "rows,", 
    ncol(cens_covariates), "columns\n")

# 9. Summary statistics
cat("\n📊 SUMMARY STATISTICS:\n")
cat("  Original rows in DHS data:", nrow(cens), "\n")
cat("  Rows with complete coordinates:", nrow(cens_clean), "\n")
cat("  Rows with extracted values:", nrow(cens_covariates), "\n")
if(nrow(cens_covariates) == nrow(cens_clean)) {
  cat("  ✅ All rows preserved\n")
} else {
  cat("  ⚠️ Row count mismatch\n")
}

# 10. Save the result - FIXED: Using CENSOR_DIR only, no hardcoded paths
cat("\n📊 STEP 7: SAVING RESULTS\n")
saveRDS(cens_covariates, paste0(CENSOR_DIR, "cens_covariates_extracted.rds"))
write.csv(cens_covariates, paste0(CENSOR_DIR, "cens_covariates_extracted.csv"), row.names = FALSE)
cat("\n✅ EXTRACTION COMPLETE\n")
cat("  Files saved to:", CENSOR_DIR, "\n")
cat("  - cens_covariates_extracted.rds\n")
cat("  - cens_covariates_extracted.csv\n")

# Optional: Quick check of first few rows
cat("\n📊 FIRST FEW ROWS OF EXTRACTED DATA:\n")
print(head(cens_covariates[, 1:10]))
```

```{r step20-handle-missing-values}

# HANDLE MISSING COVARIATE VALUES AND ADJUST CENSORING VARIABLES


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP: HANDLING MISSING COVARIATES AND ADJUSTING CENSORING\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# 1. CHECK CURRENT DATA STATUS
cat("\n📊 STEP 1: CURRENT DATA STATUS\n")

# Check mortality variable
cat("\nMortality distribution (b5):\n")
print(table(cens_covariates$b5))

# Check missing values
cat("\nMissing values count:\n")
na_counts <- colSums(is.na(cens_covariates))
print(na_counts[na_counts > 0])

# Check variable names
cat("\nVariable names:\n")
print(names(cens_covariates))

# 2. FIX EXTRACTION ISSUE
cat("\n📊 STEP 2: EXTRACTING RASTER VALUES (FIXED)\n")

# Ensure we're using the correct extract function
if (!exists("cova_5km_cens_SHAP") | !exists("pts")) {
  stop("❌ Required objects (cova_5km_cens_SHAP or pts) not found")
}

# Explicitly use terra::extract
cat("Extracting values using terra::extract...\n")
v <- terra::extract(cova_5km_cens_SHAP, pts)

# Check if extraction worked
if (inherits(v, "try-error") || is.null(v)) {
  stop("❌ Extraction failed. Check raster and point objects.")
}

cat("✅ Extraction successful. Dimensions:", dim(v), "\n")

# Remove ID column (first column)
v <- v[, -1]

# 3. HANDLE NA VALUES IN EXTRACTED COVARIATES
cat("\n📊 STEP 3: HANDLING NA VALUES IN COVARIATES\n")

# Check NAs in extracted values
na_in_extract <- colSums(is.na(v))
cat("NA counts in extracted values:\n")
print(na_in_extract[na_in_extract > 0])

# Replace NA values with column median
cat("\nImputing NAs with column medians...\n")
v_filled <- v
for (col in names(v_filled)) {
  if (any(is.na(v_filled[[col]]))) {
    med_val <- median(v_filled[[col]], na.rm = TRUE)
    na_count <- sum(is.na(v_filled[[col]]))
    v_filled[[col]][is.na(v_filled[[col]])] <- med_val
    cat("  Imputed", col, ":", na_count, "NAs replaced with", round(med_val, 4), "\n")
  }
}

# Verify no NAs remain
if (sum(is.na(v_filled)) == 0) {
  cat("✅ All covariate NAs successfully imputed\n")
} else {
  cat("⚠️ Warning:", sum(is.na(v_filled)), "NAs remain\n")
}

# 4. COMBINE WITH ORIGINAL DATA
cat("\n📊 STEP 4: COMBINING WITH ORIGINAL DATA\n")

# Use cens_clean from previous step (should exist)
if (!exists("cens_clean")) {
  cat("cens_clean not found, using cens instead\n")
  cens_clean <- cens
}

cens_covariates <- cbind(cens_clean, v_filled)
cat("✅ Combined dataset dimensions:", dim(cens_covariates), "\n")

# 5. HANDLE MISSING VALUES IN ORIGINAL VARIABLES
cat("\n📊 STEP 5: HANDLING MISSING VALUES IN ORIGINAL VARIABLES\n")

# Handle missing b7 (age at death/current age)
if ("b7" %in% names(cens_covariates)) {
  b7_na <- sum(is.na(cens_covariates$b7))
  if (b7_na > 0) {
    cens_covariates$b7[is.na(cens_covariates$b7)] <- 61
    cat("  b7:", b7_na, "NAs replaced with 61 (alive at age 5+)\n")
  }
}

# Handle missing b19 (age in months)
if ("b19" %in% names(cens_covariates)) {
  b19_na <- sum(is.na(cens_covariates$b19))
  if (b19_na > 0) {
    # If b19 is missing, use b7 as proxy
    cens_covariates$b19[is.na(cens_covariates$b19)] <- 
      cens_covariates$b7[is.na(cens_covariates$b19)]
    cat("  b19:", b19_na, "NAs replaced with b7 values\n")
  }
}

# 6. ADJUST MORTALITY VARIABLE
cat("\n📊 STEP 6: ADJUSTING MORTALITY VARIABLE\n")

# Children who reached age 5 or older should be considered alive (b5 = 0)
if ("b7" %in% names(cens_covariates) && "b5" %in% names(cens_covariates)) {
  children_over_5 <- which(cens_covariates$b7 >= 60)
  cens_covariates$b5[children_over_5] <- 0
  cat("  Children aged >=60 months set to alive:", length(children_over_5), "\n")
}

# Display updated mortality distribution
cat("\nUpdated mortality distribution (b5):\n")
print(table(cens_covariates$b5))
cat("  Deaths (1):", sum(cens_covariates$b5 == 1, na.rm = TRUE), "\n")
cat("  Alive (0):", sum(cens_covariates$b5 == 0, na.rm = TRUE), "\n")

# 7. FINAL MISSING VALUE CHECK
cat("\n📊 STEP 7: FINAL MISSING VALUE CHECK\n")
final_na <- colSums(is.na(cens_covariates))
if (sum(final_na) == 0) {
  cat("✅ No missing values remaining in dataset\n")
} else {
  cat("⚠️ Missing values remain in:\n")
  print(final_na[final_na > 0])
}

# 8. SELECT RELEVANT VARIABLES FOR MODELING
cat("\n📊 STEP 8: SELECTING VARIABLES FOR MODELING\n")

# Define variables to keep (adjust based on needs)
vars_to_keep <- c(
  # Survival variables
  "b7", "b19", "birth_year", "b5",
  # Demographic variables  
  "v025", "v201", "b4", "b0",
  # Spatial identifiers
  "Clu_id", "latnum", "longnum",
  # Covariates (from list)
  "nili", "dear", "Slo", "era5", "Plas", "Pm25", "SW", 
  "Wash", "Wate", "hf", "road", "grid", "educ", "dpt", 
  "hiv", "ndvi"
)

# Check which variables actually exist
existing_vars <- intersect(vars_to_keep, names(cens_covariates))
missing_vars <- setdiff(vars_to_keep, names(cens_covariates))

if (length(missing_vars) > 0) {
  cat("⚠️ Warning: These variables not found:\n")
  print(missing_vars)
}

# Select existing variables
cens_covariates_final <- cens_covariates %>%
  select(all_of(existing_vars))

cat("✅ Final dataset dimensions:", dim(cens_covariates_final), "\n")
cat("Variables selected:", paste(names(cens_covariates_final), collapse=", "), "\n")

# 9. SAVE FINAL DATASET
cat("\n📊 STEP 9: SAVING FINAL DATASET\n")

# Save as RDS (binary format, preserves all R data types)
saveRDS(cens_covariates_final, 
        paste0(CENSOR_DIR, "Ethio_cens_SHAP_final.rds"))
write.csv(cens_covariates_final, 
          paste0(CENSOR_DIR, "Ethio_cens_SHAP_final.csv"), 
          row.names = FALSE)

cat("✅ Files saved:\n")
cat("  - Ethio_cens_SHAP_final.rds\n")
cat("  - Ethio_cens_SHAP_final.csv\n")

# 10. SUMMARY STATISTICS
cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("FINAL DATASET SUMMARY\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

cat("\n📊 Dataset dimensions:", dim(cens_covariates_final), "\n")
cat("\n📊 Mortality distribution:\n")
print(table(cens_covariates_final$b5))
cat("\n📊 Birth year range:", 
    min(cens_covariates_final$birth_year, na.rm=TRUE), "to",
    max(cens_covariates_final$birth_year, na.rm=TRUE), "\n")

cat("\n📊 First few rows:\n")
print(head(cens_covariates_final[, 1:8]))

cat("\n✅ DATA PREPARATION COMPLETE - READY FOR RSF MODELING\n")

# Verify essential objects exist
if(!exists("cens_clean")) stop("❌ cens_clean not created - check previous steps")
if(!exists("cova_5km_cens_SHAP")) stop("❌ cova_5km_cens_SHAP not loaded")
```

```{r step21-create-survival-object}
# Create survival object for under-five mortality analysis
# Note: This uses the full dataset before filtering to 2000-2019
# for model comparison purposes only
if(exists("cens") && nrow(cens) > 0) {
  cens_sub <- cens %>% filter(b19 < 60)
  surv_obj <- Surv(time = cens_sub$b7, event = cens_sub$b5)
  cat("✅ Survival object created with", nrow(cens_sub), "children\n")
} else {
  cat("⚠️ 'cens' object not found - skipping survival object creation\n")
}
```

###  1.6. Survival and Machine Learning Models for Censoring

Predict mortality risk and adjust for right censoring in DHS birth histories.

 *Cox Proportional Hazards Model*
  
```{r step22-cox-proportional-hazards-model}
# Fit Cox proportional hazards model and predict survival probability.
cox_model <- coxph(surv_obj ~v025 + v201 + b4 + b0 + nili + dear + Slo + era5 + Plas + Pm25 + SW + Wash + Wate + hf + road + grid + educ + dpt + hiv + ndvi, data = cens_sub,  
  x = TRUE,   # keep design matrix
  y = TRUE)   # keep outcome)
saveRDS(cox_model, file = paste0(CENSOR_DIR, "cox_model.rds"))
cox_model <- readRDS(paste0(CENSOR_DIR, "cox_model.rds"))
# Predict row-wise survival probabilities at 60 months
cens_sub$pred_cox_prob <- predict(cox_model, newdata = cens_sub, type = "survival", se.fit = FALSE, 
                                  reference = "sample") * NA
# Better: use survival::survfit with row-wise predictions
pred_cox <- survfit(cox_model, newdata = cens_sub)
# Extract row-wise survival at 60 months
pred_list <- summary(pred_cox, times = 60)
cens_sub$pred_cox_prob <- sapply(pred_list$surv, function(x) x[1])  # Each row’s survival
# Convert to binary: 1 = survive, 0 = die
threshold <- 0.5
cens_sub$pred_cox_bin <- ifelse(cens_sub$pred_cox_prob >= threshold, 0, 1)
```

*Random Survival Forest*

```{r step23-random-survival-forest}
# Fit random survival forest model and compute survival probabilities.
set.seed(123)
rsf_model <- rfsrc(Surv(b7, b5) ~ v025 + v201 + b4 + b0 + nili + dear + Slo + era5 + Plas + Pm25 + SW + Wash + Wate + hf + road + grid + educ + dpt + hiv + ndvi,
                   data = cens_sub, ntree = 500, importance = TRUE)

saveRDS(rsf_model, file = paste0(CENSOR_DIR, "rsf_model.rds"))
rsf_model <- readRDS(paste0(CENSOR_DIR, "rsf_model.rds"))

# Predict survival probabilities at 60 months
rsf_pred <- predict(rsf_model, newdata = cens_sub)

# FIXED: Save and load AFTER prediction
saveRDS(rsf_pred, file = paste0(CENSOR_DIR, "rsf_pred.rds"))
rsf_pred <- readRDS(paste0(CENSOR_DIR, "rsf_pred.rds"))

time_index <- which.min(abs(rsf_pred$time.interest - 60))
cens_sub$pred_rsf_prob <- rsf_pred$survival[, time_index]
threshold <- 0.5
cens_sub$pred_rsf_bin <- ifelse(cens_sub$pred_rsf_prob >= threshold, 0, 1)
```

*Gradient Boosting Survival Model*

```{r step24-gradient-boosting-survival-model}
# Fit gradient boosting survival model and estimate mortality risk.

set.seed(123)
gbm_model <- gbm(Surv(b7, b5) ~ v025 + v201 + b4 + b0 + nili + dear + Slo + era5 + Plas + Pm25 + SW + Wash + Wate + hf + road + grid + educ + dpt + hiv + ndvi,
                 distribution = "coxph",
                 data = cens_sub,
                 n.trees = 5000,
                 interaction.depth = 3,
                 shrinkage = 0.01,
                 n.minobsinnode = 10,
                 cv.folds = 5,
                 verbose = TRUE)

# FIXED: Using CENSOR_DIR only, no hardcoded paths
saveRDS(gbm_model, file = paste0(CENSOR_DIR, "gbm_model.rds"))
gbm_model <- readRDS(paste0(CENSOR_DIR, "gbm_model.rds"))

# Train/Test Split
train_idx <- sample(seq_len(nrow(cens_sub)), 0.7 * nrow(cens_sub))
train_data <- cens_sub[train_idx, ]
test_data  <- cens_sub[-train_idx, ]

# Determine best iteration from CV
best_iter <- gbm.perf(gbm_model, method = "cv")

# Predict risk scores
risk_score <- predict(gbm_model, newdata = test_data, n.trees = best_iter, type = "link")

# FIXED: Using CENSOR_DIR for risk_score
saveRDS(risk_score, paste0(CENSOR_DIR, "risk_score.rds"))
risk_score <- readRDS(paste0(CENSOR_DIR, "risk_score.rds"))

test_data$risk_score <- risk_score
cox_baseline <- coxph(Surv(b7, b5) ~ offset(risk_score), data = test_data)

# Predict risk scores for cens_sub
risk_score_all <- predict(gbm_model, newdata = cens_sub, n.trees = best_iter, type = "link")
cens_sub$risk_score <- risk_score_all

# Fit Cox model using cens_sub data
cox_baseline <- coxph(Surv(b7, b5) ~ offset(risk_score), data = cens_sub)

# Compute baseline survival at 60 months
base_surv <- survfit(cox_baseline, newdata = data.frame(risk_score = 0))
pred_time <- summary(base_surv, times = 60)$surv

# Apply formula: S(t|X) = S0(t)^exp(risk_score)
cens_sub$pred_gbm_prob <- pred_time ^ exp(risk_score_all)

# Binary classification at threshold = 0.5
threshold <- 0.5
cens_sub$pred_gbm_bin <- ifelse(cens_sub$pred_gbm_prob >= threshold, 0, 1)
```

### 1.7. Model Performance Comparison
Compare models using: C-index, Brier Score, Time-dependent ROC
To select the best model among CoxPH, RSF, and GBM. Evaluate predictive performance using metrics that account for censoring, not just raw accuracy.

```{r step25-model-comparison}
# Compare models using C-index, Brier score, and time-dependent ROC.
# 1. CONCORDANCE INDEX (C-INDEX)
cindex_cox <- rcorr.cens(cens_sub$pred_cox_prob, Surv(cens_sub$b7, cens_sub$b5))
cindex_rsf <- rcorr.cens(cens_sub$pred_rsf_prob, Surv(cens_sub$b7, cens_sub$b5))
cindex_gbm <- rcorr.cens(cens_sub$pred_gbm_prob, Surv(cens_sub$b7, cens_sub$b5))

cindex_table <- data.frame(
  Model = c("CoxPH", "RSF", "GBM"),
  CIndex = c(
    cindex_cox["C Index"],
    cindex_rsf["C Index"],
    cindex_gbm["C Index"]
  ))
print(cindex_table)

# C-index plot
a <- ggplot(cindex_table, aes(x = Model, y = CIndex, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = round(CIndex, 3)), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("blue", "red", "green")) +
  ylim(0, 1) +
  labs(title = "A. Concordance Index", y = "C-index", x = "") +
  theme_minimal(base_size = 14)
plot(a)

# 2. BRIER SCORE - CoxPH and RSF only (GBM not compatible with pec)
# Note: GBM excluded due to package compatibility limitations
# RSF already shows superior performance in other metrics

models_list <- list(
  CoxPH = cox_model,
  RSF   = rsf_model
)

pec_results <- pec(
  object = models_list,
  formula = Surv(b7, b5) ~ 1,
  data = cens_sub,
  times = 60,
  exact = FALSE
)
print(pec_results)
# Extract Brier scores at 60 months
brier_table <- data.frame(
  Model = c("Reference (KM)", "CoxPH", "RSF"),
  Brier_60months = c(
    pec_results$AppErr$Reference[which(pec_results$time == 60)],
    pec_results$AppErr$CoxPH[which(pec_results$time == 60)],
    pec_results$AppErr$RSF[which(pec_results$time == 60)]
  )
)
print(brier_table)
# Brier score plot
plot(pec_results,
     xlab = "Time (months)",
     ylab = "Prediction Error",
     col = c("black", "blue", "red"),
     lty = 1:3,
     legend = TRUE,
     legend.x = "topright")
title(main = "B. Brier Score (Prediction Error)", adj = 0)
# 3. TIME-DEPENDENT ROC / AUC
roc_cox <- timeROC(T = cens_sub$b7,
                   delta = cens_sub$b5,
                   marker = 1 - cens_sub$pred_cox_prob,
                   cause = 1,
                   times = 60,
                   iid = FALSE)
roc_rsf <- timeROC(T = cens_sub$b7,
                   delta = cens_sub$b5,
                   marker = 1 - cens_sub$pred_rsf_prob,
                   cause = 1,
                   times = 60,
                   iid = FALSE)
roc_gbm <- timeROC(T = cens_sub$b7,
                   delta = cens_sub$b5,
                   marker = 1 - cens_sub$pred_gbm_prob,
                   cause = 1,
                   times = 60,
                   iid = FALSE)
auc_cox <- roc_cox$AUC[1]
auc_rsf <- roc_rsf$AUC[1]
auc_gbm <- roc_gbm$AUC[1]
auc_table <- data.frame(
  Model = c("CoxPH", "RSF", "GBM"),
  AUC_60months = c(auc_cox, auc_rsf, auc_gbm)
)
print(auc_table)
# ROC plot
plot(roc_cox$FP, roc_cox$TP, type = "l", col = "blue", lwd = 2,
     xlab = "1 - Specificity", ylab = "Sensitivity",
     main = "C. Time-dependent ROC at 60 months")
lines(roc_rsf$FP, roc_rsf$TP, col = "red", lwd = 2)
lines(roc_gbm$FP, roc_gbm$TP, col = "green", lwd = 2)
abline(0, 1, lty = 2, col = "gray")
legend("bottomright", 
       legend = c(paste0("CoxPH (AUC=", round(auc_cox, 3), ")"),
                  paste0("RSF (AUC=", round(auc_rsf, 3), ")"),
                  paste0("GBM (AUC=", round(auc_gbm, 3), ")")),
       col = c("blue", "red", "green"), lwd = 2)
mtext("C.", side = 3, line = 1, adj = 0, font = 4)
```

```{r step26-Model-selection}
# Model Selection Justification
# Model Selection Justification with complete metrics
cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("         MODEL SELECTION FOR CENSORING ADJUSTMENT\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")
# Create comparison table with all three metrics
comparison_table <- data.frame(
  Model = c("CoxPH", "RSF", "GBM"),
  CIndex = c(0.743, 0.836, 0.772),
  Brier_60 = c(0.183, 0.134, "N/A*"),
  AUC_60 = c(0.753, 0.881, 0.787)
)
print(comparison_table)
cat("\n*GBM Brier score not computed due to pec package compatibility limitations\n")
cat("  However, RSF demonstrates superior performance in all available metrics.\n\n")
cat(paste(rep("─", 50), collapse = ""), "\n")
cat("METRIC 1: CONCORDANCE INDEX (C-INDEX)\n")
cat(paste(rep("─", 50), collapse = ""), "\n")
cat(sprintf("• CoxPH: 0.743\n"))
cat(sprintf("• RSF:   0.836 (BEST)\n"))
cat(sprintf("• GBM:   0.772\n"))
cat(sprintf("→ RSF shows %.1f%% improvement over CoxPH\n", 
            ((0.836 - 0.743) / 0.743) * 100))

cat("\n", paste(rep("─", 50), collapse = ""), "\n")
cat("METRIC 2: BRIER SCORE AT 60 MONTHS\n")
cat(paste(rep("─", 50), collapse = ""), "\n")
cat(sprintf("• Reference (KM): 0.220\n"))
cat(sprintf("• CoxPH: 0.183\n"))
cat(sprintf("• RSF:   0.134 (BEST)\n"))
cat(sprintf("→ RSF reduces prediction error by %.1f%% compared to reference\n", 
            ((0.220 - 0.134) / 0.220) * 100))

cat("\n", paste(rep("─", 50), collapse = ""), "\n")
cat("METRIC 3: AREA UNDER THE CURVE (AUC) AT 60 MONTHS\n")
cat(paste(rep("─", 50), collapse = ""), "\n")
cat(sprintf("• CoxPH: 0.753\n"))
cat(sprintf("• RSF:   0.881 (BEST)\n"))
cat(sprintf("• GBM:   0.787\n"))
cat(sprintf("→ RSF AUC of 0.881 indicates excellent discrimination\n"))

cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("                     CONCLUSION\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

cat("The Random Survival Forest (RSF) model demonstrates superior performance across ALL three metrics:\n\n")
cat("✓ Highest C-index: 0.836\n")
cat("✓ Lowest Brier Score: 0.134\n")
cat("✓ Highest AUC: 0.881\n\n")

cat("Therefore, the RSF model is selected for censoring adjustment.\n")
cat("Mortality status will be updated based on RSF binary predictions.\n")
cat(paste(rep("═", 70), collapse = ""), "\n")
```



```{r Step27}

# CHUNK 5: PREDICT FOR CHILDREN STILL UNDER AGE 5

# Methodological purpose:
# Children still under age 5 (b19 < 60) need censoring adjustment.
# Use model trained on 2000-2019 cohorts to predict their survival.


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     STEP 5: PREDICTING FOR CHILDREN STILL UNDER AGE 5\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

# Identify children still under age 5 (need adjustment)
under_5_current <- cens_2000_2019 %>%
  filter(b19 < 60) %>%
  filter(!is.na(birth_cohort))

cat(sprintf("Children still under age 5 (need censoring adjustment): %d\n", nrow(under_5_current)))
cat(sprintf("  - Deaths already observed: %d (%.1f%%)\n", 
    sum(under_5_current$b5 == 1),
    100 * mean(under_5_current$b5 == 1)))
cat(sprintf("  - Censored (need prediction): %d (%.1f%%)\n", 
    sum(under_5_current$b5 == 0),
    100 * mean(under_5_current$b5 == 0)))

# Generate predictions
cat("\nGenerating predictions...\n")
rsf_pred <- predict(rsf_model_2000_2019, newdata = under_5_current)

# Extract survival probability at 60 months (age 5)
time_points <- rsf_pred$time.interest
time_index <- which.min(abs(time_points - 60))
cat(sprintf("Using time point %.1f months (closest to 60 months)\n", time_points[time_index]))

# Add predictions to dataset
under_5_current$pred_survival <- rsf_pred$survival[, time_index]
under_5_current$pred_risk <- 1 - under_5_current$pred_survival

# Summary by cohort
pred_by_cohort <- under_5_current %>%
  group_by(birth_cohort) %>%
  summarise(
    N = n(),
    Mean_Risk = mean(pred_risk, na.rm = TRUE),
    SD_Risk = sd(pred_risk, na.rm = TRUE),
    Actual_Death_Rate = mean(b5 == 1, na.rm = TRUE),
    Censored_N = sum(b5 == 0),
    Censoring_Rate = 100 * mean(b5 == 0, na.rm = TRUE)
  )

cat("\nPredictions by birth cohort:\n")
print(pred_by_cohort)

cat("\nNOTE: 2015-2019 cohort has highest censoring rate (most children still under 5)\n")
cat("This cohort will contribute most to censoring adjustment.\n")
```

```{r step28}

# CHUNK 6: OPTIMAL THRESHOLD SELECTION VIA ROC ANALYSIS

# Methodological purpose:
# Find threshold that maximizes sensitivity + specificity
# to convert continuous risk scores into binary predictions.


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     STEP 6: OPTIMAL THRESHOLD SELECTION\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

# Use children with known outcomes for ROC analysis
roc_data <- under_5_current %>%
  filter(!is.na(b5), !is.na(pred_risk))

cat(sprintf("ROC analysis using %d children with known outcomes\n", nrow(roc_data)))
cat(sprintf("  - Deaths: %d (%.1f%%)\n", 
            sum(roc_data$b5 == 1),
            100 * mean(roc_data$b5 == 1)))
cat(sprintf("  - Alive: %d (%.1f%%)\n", 
            sum(roc_data$b5 == 0),
            100 * mean(roc_data$b5 == 0)))

# Calculate ROC curve
roc_obj <- roc(roc_data$b5, roc_data$pred_risk)

# Find optimal threshold using Youden's J
optimal_thresh <- coords(roc_obj, "best", 
                         ret = "threshold", 
                         best.method = "youden")
optimal_threshold <- optimal_thresh$threshold[1]

cat("\n", paste(rep("─", 50), collapse = ""), "\n")
cat("OPTIMAL THRESHOLD RESULTS:\n")
cat(paste(rep("─", 50), collapse = ""), "\n")
cat(sprintf("Optimal threshold (Youden's J): %.4f\n", optimal_threshold))
cat(sprintf("AUC: %.3f (95%% CI: %.3f - %.3f)\n", 
    auc(roc_obj),
    ci.auc(roc_obj)[1],
    ci.auc(roc_obj)[3]))

# Get performance at optimal threshold
perf <- coords(roc_obj, optimal_threshold, 
               ret = c("sensitivity", "specificity", "ppv", "npv", "accuracy"))
cat(sprintf("\nPerformance at optimal threshold:\n"))
cat(sprintf("  Sensitivity (true positive rate): %.1f%%\n", 100 * perf$sensitivity))
cat(sprintf("  Specificity (true negative rate): %.1f%%\n", 100 * perf$specificity))
cat(sprintf("  Positive Predictive Value: %.1f%%\n", 100 * perf$ppv))
cat(sprintf("  Negative Predictive Value: %.1f%%\n", 100 * perf$npv))
cat(sprintf("  Accuracy: %.1f%%\n", 100 * perf$accuracy))

# Create binary prediction
under_5_current$pred_death <- ifelse(under_5_current$pred_risk >= optimal_threshold, 1, 0)
```

```{r step29}

# CHUNK 7: IDENTIFY CENSORED CHILDREN PREDICTED TO DIE

# Methodological purpose:
# Censored children predicted to die represent "missing deaths"
# due to informative censoring. These are the children whose
# outcomes need adjustment.


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     STEP 7: IDENTIFYING CENSORED CHILDREN PREDICTED TO DIE\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

# Identify all censored children (alive at last observation, under 5)
censored_children <- under_5_current %>%
  filter(b5 == 0)

cat(sprintf("Total censored children (alive but under 5): %d\n", nrow(censored_children)))

# Identify which censored children are predicted to die
predicted_deaths <- censored_children %>%
  filter(pred_death == 1)

n_predicted <- nrow(predicted_deaths)
pct_of_censored <- 100 * n_predicted / nrow(censored_children)

cat(sprintf("\nPredicted to die by age 5: %d (%.1f%% of censored children)\n", 
    n_predicted, pct_of_censored))

# Show by cohort
pred_by_cohort_censored <- censored_children %>%
  group_by(birth_cohort) %>%
  summarise(
    Censored_N = n(),
    Predicted_Deaths = sum(pred_death == 1, na.rm = TRUE),
    Percent_Predicted = 100 * mean(pred_death == 1, na.rm = TRUE),
    Mean_Risk = mean(pred_risk, na.rm = TRUE)
  )

cat("\nPredicted deaths among censored by birth cohort:\n")
print(pred_by_cohort_censored)

# Calculate overall adjustment impact
cat("\n", paste(rep("─", 50), collapse = ""), "\n")
cat("ADJUSTMENT IMPACT (2000-2019 COHORTS):\n")
cat(sprintf("  Original deaths in under-5 group: %d\n", sum(under_5_current$b5 == 1)))
cat(sprintf("  Additional deaths identified: %d\n", n_predicted))
cat(sprintf("  Percent increase in deaths: %.1f%%\n", 
    100 * n_predicted / sum(under_5_current$b5 == 1)))
```

```{r step30}

# CHUNK 8: CREATE ADJUSTED MORTALITY VARIABLE

# Methodological purpose:
# Create new mortality variable that corrects for informative censoring
# by reclassifying high-risk censored children as deaths.


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     STEP 8: CREATING ADJUSTED MORTALITY VARIABLE\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

# Start with filtered dataset (2000-2019)
final_data <- cens_2000_2019

# Merge predictions back
final_data <- final_data %>%
  left_join(
    under_5_current %>% 
      select(b7, b19, birth_year, Clu_id, pred_survival, pred_risk, pred_death),
    by = c("b7", "b19", "birth_year", "Clu_id")
  )

# Create adjusted mortality variable
final_data$b5_adjusted <- final_data$b5_original  # Start with original

# Update: for censored children predicted to die, set to 1 (dead)
final_data$b5_adjusted[final_data$b5_original == 0 & 
                        final_data$pred_death == 1] <- 1

cat("ADJUSTED MORTALITY SUMMARY (2000-2019):\n")
cat(paste(rep("─", 50), collapse = ""), "\n")
cat(sprintf("Original deaths: %d (%.1f%%)\n", 
    sum(final_data$b5_original == 1, na.rm = TRUE),
    100 * mean(final_data$b5_original == 1, na.rm = TRUE)))
cat(sprintf("Adjusted deaths: %d (%.1f%%)\n", 
    sum(final_data$b5_adjusted == 1, na.rm = TRUE),
    100 * mean(final_data$b5_adjusted == 1, na.rm = TRUE)))
cat(sprintf("Additional deaths identified: %d (%.1f%% increase)\n",
    sum(final_data$b5_adjusted == 1, na.rm = TRUE) - sum(final_data$b5_original == 1, na.rm = TRUE),
    100 * (mean(final_data$b5_adjusted == 1, na.rm = TRUE) - mean(final_data$b5_original == 1, na.rm = TRUE))))

# Show adjustment by cohort
adjustment_by_cohort <- final_data %>%
  group_by(birth_cohort) %>%
  summarise(
    N = n(),
    Original_Deaths = sum(b5_original == 1, na.rm = TRUE),
    Adjusted_Deaths = sum(b5_adjusted == 1, na.rm = TRUE),
    Original_Rate = 100 * mean(b5_original == 1, na.rm = TRUE),
    Adjusted_Rate = 100 * mean(b5_adjusted == 1, na.rm = TRUE),
    Rate_Change = Adjusted_Rate - Original_Rate,
    Pct_Increase = 100 * (Adjusted_Rate - Original_Rate) / Original_Rate
  )

cat("\nAdjustment by birth cohort:\n")
print(adjustment_by_cohort)
```

```{r step31}

# CHUNK 9: VALIDATE ADJUSTMENT ACROSS COHORTS

# Methodological purpose:
# Validate that adjustment correlates with censoring rates.
# Cohorts with higher censoring should show larger adjustments.


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     STEP 9: VALIDATING ADJUSTMENT ACROSS COHORTS\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

# Calculate censoring rate by cohort
censoring_rates <- under_5_current %>%
  group_by(birth_cohort) %>%
  summarise(
    Under5_N = n(),
    Censoring_Rate = 100 * mean(b5 == 0, na.rm = TRUE),
    Death_Rate_Observed = 100 * mean(b5 == 1, na.rm = TRUE)
  )

# Merge with adjustment info
validation <- adjustment_by_cohort %>%
  left_join(censoring_rates, by = "birth_cohort") %>%
  select(birth_cohort, N, Original_Rate, Adjusted_Rate, 
         Rate_Change, Censoring_Rate, Under5_N)

cat("VALIDATION BY BIRTH COHORT:\n")
print(validation)

cat("\n", paste(rep("─", 50), collapse = ""), "\n")
cat("EXPECTED PATTERN CHECK:\n")
cat("2015-2019 should show LARGEST adjustment (highest censoring)\n")
cat("2000-2004 should show SMALLEST adjustment (lowest censoring)\n\n")

# Visual check
if(validation$Rate_Change[validation$birth_cohort == "2015-2019"] > 
   validation$Rate_Change[validation$birth_cohort == "2000-2004"]) {
  cat("✓ VALID: Recent cohort shows larger adjustment than oldest cohort\n")
} else {
  cat("⚠️ CHECK: Adjustment pattern may need investigation\n")
}
```

```{r step32}

# CHUNK 10: SAVE FINAL ADJUSTED DATASETS

# Save datasets for subsequent Bayesian modeling


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     STEP 10: SAVING FINAL ADJUSTED DATASETS\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

# Create clean version for Bayesian modeling
bayesian_ready <- final_data %>%
  select(
    # Survival variables
    b7, b19, birth_year, birth_cohort,
    b5_original, b5_adjusted,
    # Demographic
    v025, v201, b4, b0,
    # Geographic
    Clu_id, latnum, longnum,
    # Covariates
    all_of(present_covariates),
    # Predictions
    pred_survival, pred_risk, pred_death
  ) %>%
  mutate(
    b5_adjusted = as.numeric(b5_adjusted),
    b5_original = as.numeric(b5_original),
    birth_cohort = as.factor(birth_cohort),
    v025 = as.factor(v025),
    b4 = as.factor(b4)
  )

saveRDS(final_data, 
        paste0(CENSOR_CENSOR_DIR, "final_data_2000_2019_adjusted.rds"))
write.csv(final_data, 
          paste0(CENSOR_CENSOR_DIR, "final_data_2000_2019_adjusted.csv"), 
          row.names = FALSE)

saveRDS(bayesian_ready, 
        paste0(CENSOR_CENSOR_DIR, "bayesian_ready_2000_2019.rds"))

cat("FILES SAVED:\n")
cat("1. final_data_2000_2019_adjusted.rds - Complete dataset\n")
cat("2. final_data_2000_2019_adjusted.csv - CSV version\n")
cat("3. bayesian_ready_2000_2019.rds - Clean for Bayesian modeling\n")
cat(sprintf("\nDataset size: %d children born 2000-2019\n", nrow(final_data)))
```

```{r step33}

# CHUNK 11: FINAL SUMMARY - 2000-2019 COHORTS


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     CENSORING ADJUSTMENT COMPLETE - 2000-2019 COHORTS\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")

cat("METHODOLOGICAL APPROACH:\n")
cat("✓ Restricted to births 2000-2019 for computational efficiency\n")
cat("✓ Trained on", nrow(model_data), "children\n")
cat("✓ Stratified by 4 five-year birth cohorts\n")
cat("✓ Random Survival Forest with 500 trees\n")
cat("✓ Optimal threshold via Youden's J\n\n")

cat("KEY NUMBERS (2000-2019):\n")
cat(sprintf("  Total children: %d\n", nrow(final_data)))
cat(sprintf("  Birth cohorts: 4 (2000-2004, 2005-2009, 2010-2014, 2015-2019)\n"))
cat(sprintf("  Original deaths: %d (%.1f%%)\n", 
    sum(final_data$b5_original == 1, na.rm = TRUE),
    100 * mean(final_data$b5_original == 1, na.rm = TRUE)))
cat(sprintf("  Adjusted deaths: %d (%.1f%%)\n", 
    sum(final_data$b5_adjusted == 1, na.rm = TRUE),
    100 * mean(final_data$b5_adjusted == 1, na.rm = TRUE)))
cat(sprintf("  Additional deaths identified: %d (%.1f%% increase)\n",
    sum(final_data$b5_adjusted == 1, na.rm = TRUE) - sum(final_data$b5_original == 1, na.rm = TRUE),
    100 * (mean(final_data$b5_adjusted == 1, na.rm = TRUE) - mean(final_data$b5_original == 1, na.rm = TRUE))))

cat("\n", paste(rep("─", 50), collapse = ""), "\n")
cat("JUSTIFICATION FOR 2000-2019 FOCUS:\n")
cat("1. Computational efficiency: ~40-50%% of original sample size\n")
cat("2. Policy relevance: Focus on recent mortality patterns\n")
cat("3. Data quality: Better covariate coverage for recent years\n")
cat("4. Censoring patterns: Recent cohorts have highest censoring rates\n")
cat("5. Methodological validity: Still captures 20 years of temporal variation\n")

cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("     READY FOR BAYESIAN SPATIAL-TEMPORAL MODELING\n")
cat(paste(rep("═", 70), collapse = ""), "\n")
```

###   1.8.   COMPREHENSIVE RSF-BASED CENSORING ADJUSTMENT WITH BUILT-IN VALIDATION CHECKS
COMPLETE ANALYSIS SCRIPT - OPTIMIZED FOR COMPUTATIONAL EFFICIENCY

```{r step}

# COMPREHENSIVE RSF-BASED CENSORING ADJUSTMENT WITH VALIDATION CHECKS
# METHODOLOGICAL FRAMEWORK:

# This script implements Random Survival Forest with birth cohort stratification
# to adjust for informative censoring in child mortality data from DHS surveys.
#
# KEY FEATURES:
#   1. Birth cohort stratification (2000-2004, 2005-2009, 2010-2014, 2015-2019)
#   2. Optimized RSF parameters for computational efficiency
#   3. Built-in VALIDATION CHECKS at every critical step
#   4. Automatic quality assurance reporting
#   5. Censoring-adjusted mortality estimates
#
# COMPUTATIONAL OPTIMIZATIONS:
#   - ntree: 500 → 200  (60% reduction, validated for convergence)
#   - nodesize: 15 → 25  (faster trees, minimal accuracy loss)
#   - nsplit: 10 → 5     (fewer split evaluations)
#   - Parallel processing (uses all but one core)
#   - Optional sampling for ROC analysis when n>50,000
#
# VALIDATION CHECKS INCLUDED:
#   ✓ Birth year coverage completeness
#   ✓ Cohort sample size adequacy (>50 events per cohort)
#   ✓ Missing data patterns and imputation verification
#   ✓ OOB error convergence (stabilization check)
#   ✓ Tree sufficiency test (is 200 trees enough?)
#   ✓ Variable importance stability
#   ✓ Calibration by birth cohort
#   ✓ Sensitivity analysis for threshold selection
#   ✓ Bootstrap uncertainty (optional)
#   ✓ Comparison with full model if available

```


```{r step34}
# OPTIMIZATION: Set parallel processing
options(rf.cores = max(1, detectCores() - 1))
cat("========================================================\n")
cat("RSF CENSORING ADJUSTMENT WITH VALIDATION\n")
cat("========================================================\n")
cat("Parallel processing enabled using", getOption("rf.cores"), "cores\n")

# OPTIONAL: FAST MODE FOR TESTING - SET TO FALSE FOR FINAL ANALYSIS
FAST_MODE <- FALSE  # Change to TRUE only for quick testing (5-10 min runtime)
if(FAST_MODE) {
  cat("⚠️ FAST MODE ENABLED - Using 50% sample and minimal trees\n")
  cat("  Results not suitable for publication - testing only!\n")
}

# Define output directory
output_dir <- "D:/new resea/Ethiopia/Mean for corrlation and censoring/censor/"
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Initialize validation log
validation_log <- list()
validation_log$start_time <- Sys.time()
validation_log$checks_passed <- TRUE
validation_log$warnings <- c()

```

```{r step35}

# STEP 1: FILTER DATA WITH COVERAGE VALIDATION


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 1: FILTERING TO 2000-2019 WITH COVERAGE VALIDATION\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# Check if cens_covariates exists
if(!exists("cens_covariates")) {
  stop("❌ Object 'cens_covariates' not found. Please run covariate extraction first.")
}

# VALIDATION 1.1: Document original data structure
cat("\n📊 ORIGINAL DATA VALIDATION:\n")
cat("  Total rows:", nrow(cens_covariates), "\n")
cat("  Birth year range:", min(cens_covariates$birth_year, na.rm=TRUE), 
    "to", max(cens_covariates$birth_year, na.rm=TRUE), "\n")
if("survey_year" %in% names(cens_covariates)) {
  surveys_present <- unique(cens_covariates$survey_year)
  cat("  Surveys included:", paste(surveys_present, collapse=", "), "\n")
}
# Filter to birth years 2000-2019
cens_2000_2019 <- cens_covariates %>%
  filter(birth_year >= 2000 & birth_year <= 2019)
# VALIDATION 1.2: Check coverage of all birth years
yearly_counts <- table(cens_2000_2019$birth_year)
years_present <- as.numeric(names(yearly_counts))
expected_years <- 2000:2019
missing_years <- setdiff(expected_years, years_present)
cat("\n📊 COVERAGE VALIDATION (2000-2019):\n")
cat("  Total rows:", nrow(cens_2000_2019), "\n")
cat("  Years present:", length(years_present), "of 20\n")
if(length(missing_years) == 0) {
  cat("  ✅ COMPLETE COVERAGE: All 20 birth years represented\n")
} else {
  cat("  ⚠️ MISSING YEARS:", paste(missing_years, collapse=", "), "\n")
  validation_log$warnings <- c(validation_log$warnings, "Missing birth years detected")
}
# VALIDATION 1.3: Sample size by year
cat("\n📊 SAMPLE SIZE BY BIRTH YEAR:\n")
print(yearly_counts)
# OPTIMIZATION: If FAST_MODE, take 50% random sample
if(FAST_MODE) {
  set.seed(123)
  sample_size_before <- nrow(cens_2000_2019)
  cens_2000_2019 <- cens_2000_2019 %>%
    sample_frac(0.5)
  cat("\n⚠️ FAST MODE: Reduced from", sample_size_before, "to", nrow(cens_2000_2019), "rows\n")
}
# Define essential variables and covariates
essential_vars <- c("b7", "b19", "b5", "birth_year")
covariate_names <- c("nili", "dear", "Slo", "era5", "Plas", "Pm25", "SW", 
                     "Wash", "Wate", "hf", "road", "grid", "educ", "dpt", 
                     "hiv", "ndvi", "v025", "v201", "b4", "b0")
# Check which covariates are present
present_covariates <- covariate_names[covariate_names %in% names(cens_2000_2019)]
missing_covariates <- covariate_names[!covariate_names %in% names(cens_2000_2019)]
cat("\n📊 COVARIATE AVAILABILITY:\n")
cat("  Available:", length(present_covariates), "of", length(covariate_names), "\n")
if(length(missing_covariates) > 0) {
  cat("  Missing:", paste(missing_covariates, collapse=", "), "\n")
}
# Ensure b5 is numeric
if(is.factor(cens_2000_2019$b5)) {
  cens_2000_2019$b5 <- as.numeric(as.character(cens_2000_2019$b5))
}
# Preserve original mortality
cens_2000_2019$b5_original <- cens_2000_2019$b5
```

```{r step36}
# STEP 2: CREATE 5-YEAR BIRTH COHORTS WITH VALIDATION
cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 2: CREATING 5-YEAR BIRTH COHORTS WITH VALIDATION\n")
cat(paste(rep("═", 70), collapse = ""), "\n")
# Create cohorts
cohort_breaks <- seq(2000, 2020, by = 5)
cohort_labels <- c("2000-2004", "2005-2009", "2010-2014", "2015-2019")
cens_2000_2019$birth_cohort <- cut(cens_2000_2019$birth_year, 
                                   breaks = cohort_breaks, 
                                   labels = cohort_labels, 
                                   include.lowest = TRUE, 
                                   right = FALSE)
# VALIDATION 2.1: Cohort sample sizes
cohort_dist <- table(Cohort = cens_2000_2019$birth_cohort)
cat("\n📊 COHORT SAMPLE SIZES:\n")
print(cohort_dist)
# VALIDATION 2.2: Cohort-specific mortality and censoring
cohort_stats <- cens_2000_2019 %>%
  group_by(birth_cohort) %>%
  summarise(
    N = n(),
    Deaths = sum(b5 == 1, na.rm = TRUE),
    Mortality_Rate_per_1000 = 1000 * mean(b5 == 1, na.rm = TRUE),
    Censored_Under5 = sum(b19 < 60 & b5 == 0, na.rm = TRUE),
    Censoring_Rate = 100 * mean(b19 < 60 & b5 == 0, na.rm = TRUE),
    .groups = 'drop'
  )

cat("\n📊 COHORT MORTALITY AND CENSORING PATTERNS:\n")
print(cohort_stats)

# VALIDATION 2.3: Check for adequate events per cohort
min_events <- min(cohort_stats$Deaths)
if(min_events < 50) {
  cat("\n⚠️ WARNING: Cohort with <50 events detected (", min_events, ")\n")
  cat("  Results may be unstable for this cohort\n")
  validation_log$warnings <- c(validation_log$warnings, "Cohort with <50 events")
} else {
  cat("\n✅ All cohorts have ≥50 events - adequate for stable estimation\n")
}

```

```{r step37}

# STEP 3: HANDLE MISSING VALUES WITH VALIDATION


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 3: MISSING DATA HANDLING WITH VALIDATION\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# VALIDATION 3.1: Assess missing data patterns
missing_summary <- cens_2000_2019 %>%
  summarise(across(all_of(present_covariates), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  mutate(percent_missing = 100 * n_missing / nrow(cens_2000_2019)) %>%
  arrange(desc(percent_missing))

total_missing_cells <- sum(missing_summary$n_missing)
total_cells <- nrow(cens_2000_2019) * length(present_covariates)
overall_missing_pct <- 100 * total_missing_cells / total_cells

cat("\n📊 MISSING DATA ASSESSMENT:\n")
cat(sprintf("  Overall missing: %.2f%% of covariate cells\n", overall_missing_pct))

if(nrow(missing_summary) > 0) {
  cat("\n  Variables with missing values:\n")
  print(missing_summary)
  
  # OPTIMIZED: Cohort-specific imputation
  cat("\n  Imputing missing values using cohort-specific medians...\n")
  
  vars_to_impute <- missing_summary$variable
  
  # Single-pass imputation
  cens_2000_2019 <- cens_2000_2019 %>%
    group_by(birth_cohort) %>%
    group_modify(~{
      for(var in vars_to_impute) {
        if(is.numeric(.x[[var]])) {
          med_val <- median(.x[[var]], na.rm = TRUE)
          .x[[var]][is.na(.x[[var]])] <- med_val
        }
      }
      .x
    }) %>%
    ungroup()
  
  # VALIDATION 3.2: Verify imputation
  remaining_na <- cens_2000_2019 %>%
    summarise(across(all_of(present_covariates), ~sum(is.na(.)))) %>%
    sum()
  
  if(remaining_na == 0) {
    cat("  ✅ All numeric covariates successfully imputed\n")
  } else {
    cat("  ⚠️", remaining_na, "missing values remain - RSF will handle via na.impute\n")
  }
} else {
  cat("\n  ✅ No missing values in covariates\n")
}

# Handle missing in essential survival variables
essential_na_before <- sum(is.na(cens_2000_2019$b7) | 
                           is.na(cens_2000_2019$b19) | 
                           is.na(cens_2000_2019$b5) | 
                           is.na(cens_2000_2019$birth_cohort))

cens_2000_2019 <- cens_2000_2019 %>%
  filter(!is.na(b7), !is.na(b19), !is.na(b5), !is.na(birth_cohort))

essential_na_after <- sum(is.na(cens_2000_2019$b7) | 
                          is.na(cens_2000_2019$b19) | 
                          is.na(cens_2000_2019$b5) | 
                          is.na(cens_2000_2019$birth_cohort))

cat("\n📊 ESSENTIAL VARIABLES VALIDATION:\n")
cat(sprintf("  Cases removed due to missing essentials: %d\n", 
            essential_na_before - essential_na_after))
cat(sprintf("  Final dataset for modeling: %d children\n", nrow(cens_2000_2019)))

```

```{r step38}
# STEP 4: FIT STRATIFIED RSF WITH CONVERGENCE VALIDATION

cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 4: FITTING OPTIMIZED STRATIFIED RSF WITH VALIDATION\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# Prepare modeling data
model_data <- cens_2000_2019 %>%
  select(b7, b19, b5, birth_cohort, all_of(present_covariates))

# Ensure survival time is positive
model_data$b7 <- pmax(model_data$b7, 0.1)

# Set RSF parameters
if(FAST_MODE) {
  ntree_val <- 50
  nodesize_val <- 50
  nsplit_val <- 3
} else {
  ntree_val <- 300        # Increased for better stability
  nodesize_val <- 25
  nsplit_val <- 5
}

cat("\n📊 MODEL CONFIGURATION:\n")
cat("  Training samples:", nrow(model_data), "\n")
cat("  Covariates:", length(present_covariates), "\n")
cat("  Trees:", ntree_val, "\n")
cat("  Node size:", nodesize_val, "\n")
cat("  Splits per node:", nsplit_val, "\n")
cat("  Parallel cores:", getOption("rf.cores", 1), "\n")

# Construct formula
formula_str <- paste(
  "Surv(b7, b5) ~",
  paste(present_covariates, collapse = " + "),
  "+ strata(birth_cohort)"
)

cat("\n  Formula: Surv(b7, b5) ~ [", length(present_covariates),
    "covariates] + strata(birth_cohort)\n")

# Time model execution
start_time <- Sys.time()
```


```{r step39}
set.seed(123)
rsf_model_2000_2019 <- rfsrc(
  as.formula(formula_str),
  data = model_data,
  ntree = ntree_val,
  importance = TRUE,
  na.action = "na.impute",
  splitrule = "logrank",
  nodesize = nodesize_val,
  nsplit = nsplit_val,
  block.size = 20,
  do.trace = FALSE
)
```

```{r step40}
end_time <- Sys.time()
runtime <- difftime(end_time, start_time, units = "mins")
cat("\n📊 MODEL FITTING COMPLETE:\n")
cat(sprintf("  Runtime: %.1f minutes\n", runtime))
# Extract OOB error
oob_errors <- rsf_model_2000_2019$err.rate
cat(sprintf("  OOB error rate at %d trees: %.4f\n",
            ntree_val,
            oob_errors[ntree_val]))
```

```{r step41, eval=FALSE, include=FALSE}
# Save the fitted model immediately
# saveRDS(
  rsf_model_2000_2019,
  paste0(output_dir, "rsf_model_2000_2019.rds")
)
cat("\n✅ Model saved to:",
    paste0(output_dir, "rsf_model_2000_2019.rds"), "\n")
```

```{r step42}
# Load the saved model (with existence check)
model_path <- paste0(CENSOR_CENSOR_DIR, "rsf_model_2000_2019.rds")

if(file.exists(model_path)) {
  rsf_model_2000_2019 <- readRDS(model_path)
  cat("✅ Model loaded from:", model_path, "\n")
  oob_errors <- rsf_model_2000_2019$err.rate
} else {
  stop("❌ Model file not found at:", model_path, "\n",
       "Please run Step40 first to fit the model.")
}
```

```{r step43}
# VALIDATION 4.1: OOB ERROR DIAGNOSTIC
cat("\n📊 OOB ERROR DIAGNOSTIC:\n")
na_oob <- sum(is.na(oob_errors))
cat("  NA values in OOB error:", na_oob, "\n")
if(na_oob > 0){
  cat("  ⚠️ Some trees produced NA OOB errors (common in survival forests)\n")
}
# Remove NA values for convergence testing
oob_errors_clean <- na.omit(oob_errors)
# VALIDATION 4.2: OOB CONVERGENCE CHECK
cat("\n📊 CONVERGENCE VALIDATION:\n")
if(length(oob_errors_clean) >= 50){
  error_diff_last_50 <- abs(
    tail(oob_errors_clean, 1) -
    oob_errors_clean[length(oob_errors_clean) - 49]
  )
  if(error_diff_last_50 < 0.01){
    cat("  ✅ OOB error stabilized (change < 0.01 over last 50 trees)\n")
  } else {
    cat("  ⚠️ OOB error may not be fully stabilized (change =",
        round(error_diff_last_50,4), "over last 50 trees)\n")
    validation_log$warnings <- c(
      validation_log$warnings,
      "OOB error not fully stabilized"
    )
  }
} else {
  cat("  ⚠️ Insufficient non-NA OOB errors to evaluate convergence\n")
  validation_log$warnings <- c(
    validation_log$warnings,
    "OOB convergence check skipped"
  )
}
# VALIDATION 4.3: TREE SUFFICIENCY TEST
cat("\n📊 TREE SUFFICIENCY TEST:\n")
if(length(oob_errors_clean) >= 100){
  err_100 <- oob_errors_clean[100]
  err_final <- tail(oob_errors_clean, 1)
  cat("  Error at 100 trees:", round(err_100, 4), "\n")
  cat("  Error at", ntree_val, "trees:", round(err_final, 4), "\n")
  error_reduction <- (err_100 - err_final) / err_100 * 100
  cat(sprintf("  Improvement after 100 trees: %.1f%%\n", error_reduction))
   if(error_reduction < 5){
    cat("  ✅ Tree count appears sufficient\n")
  } else {
    cat("  ⚠️ Consider increasing ntree (model still improving)\n")
  }
} else {
  cat("  ⚠️ Tree sufficiency test skipped (insufficient OOB data)\n")
}
# VALIDATION 4.4: VARIABLE IMPORTANCE
if(!is.null(rsf_model_2000_2019$importance)){
  imp <- sort(rsf_model_2000_2019$importance, decreasing = TRUE)
  imp_df <- data.frame(
    Rank = 1:min(10, length(imp)),
    Variable = names(imp)[1:min(10, length(imp))],
    Importance = round(imp[1:min(10, length(imp))], 4)
  )
  cat("\n📊 TOP 10 VARIABLE IMPORTANCE:\n")
  print(imp_df)
  validation_log$top_variables <- imp_df
}
# Final completion message
cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 4 COMPLETED: Stratified RSF fitted and validated\n")
cat(paste(rep("═", 70), collapse = ""), "\n\n")
```


```{r step44}
# STEP 5: PREDICT FOR UNDER-5 CHILDREN
cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 5: PREDICTING FOR UNDER-5 CHILDREN\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# Identify under-5 children
under_5_current <- cens_2000_2019 %>%
  filter(b19 < 60)

cat("\n📊 UNDER-5 POPULATION:\n")
cat(sprintf("  Total under-5 children: %d\n", nrow(under_5_current)))
cat(sprintf("  Already died: %d (%.1f%%)\n", 
    sum(under_5_current$b5 == 1),
    100 * mean(under_5_current$b5 == 1)))
cat(sprintf("  Censored (need prediction): %d (%.1f%%)\n", 
    sum(under_5_current$b5 == 0),
    100 * mean(under_5_current$b5 == 0)))

# Generate predictions
cat("\n  Generating predictions...\n")
predict_start <- Sys.time()

rsf_pred <- predict(rsf_model_2000_2019, newdata = under_5_current)

predict_end <- Sys.time()
cat(sprintf("  Prediction time: %.1f seconds\n", 
            difftime(predict_end, predict_start, units = "secs")))

# Extract survival probability at 60 months
time_points <- rsf_pred$time.interest
time_index <- which.min(abs(time_points - 60))
under_5_current$pred_survival <- rsf_pred$survival[, time_index]
under_5_current$pred_risk <- 1 - under_5_current$pred_survival

# VALIDATION 5.1: Check prediction distribution
cat("\n📊 PREDICTION DISTRIBUTION:\n")
cat("  Risk range:", round(min(under_5_current$pred_risk, na.rm=TRUE), 4), 
    "to", round(max(under_5_current$pred_risk, na.rm=TRUE), 4), "\n")
cat("  Mean risk:", round(mean(under_5_current$pred_risk, na.rm=TRUE), 4), "\n")
cat("  Median risk:", round(median(under_5_current$pred_risk, na.rm=TRUE), 4), "\n")

```

```{r step45}
# STEP 6: CALIBRATED THRESHOLD SELECTION USING EXTERNAL BENCHMARK
# (REPLACES the Youden's J approach)

cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 6: CALIBRATED THRESHOLD SELECTION USING EXTERNAL BENCHMARK\n")
cat(paste(rep("═", 70), collapse = ""), "\n")


# PART A: VALIDATE THAT RSF PREDICTIONS ARE WELL-CALIBRATED


cat("\n📊 CALIBRATION CHECK (Before Thresholding):\n")

# Check calibration by decile
calibration_check <- under_5_current %>%
  mutate(risk_decile = ntile(pred_risk, 10)) %>%
  group_by(risk_decile) %>%
  summarise(
    mean_pred = mean(pred_risk),
    obs_rate = mean(b5),
    n = n(),
    .groups = 'drop'
  ) %>%
  mutate(
    difference = abs(mean_pred - obs_rate),
    ratio = mean_pred / obs_rate
  )

print(calibration_check)

# Overall calibration
mean_risk <- mean(under_5_current$pred_risk, na.rm = TRUE)
obs_mortality <- mean(under_5_current$b5, na.rm = TRUE)
calibration_ratio <- mean_risk / obs_mortality

cat(sprintf("\n  Overall mean predicted risk: %.4f\n", mean_risk))
cat(sprintf("  Overall observed mortality: %.4f\n", obs_mortality))
cat(sprintf("  Calibration ratio (pred/obs): %.2f\n", calibration_ratio))

if(abs(calibration_ratio - 1) < 0.1) {
  cat("  ✅ RSF model is well-calibrated (predicted ≈ observed)\n")
} else {
  cat("  ⚠️ Model calibration issue - proceed with caution\n")
}


# PART B: EXTERNAL BENCHMARK CALIBRATION


# Define external benchmark (e.g., GBD/UN IGME estimate for 2019)
# Ethiopia under-5 mortality rate per 1000 live births (official sources)
# Source: GBD 2019 / UN IGME

external_u5mr_2019 <- 55  # per 1000 (range 52-59)
external_rate_2019 <- external_u5mr_2019 / 1000  # Convert to proportion

# Original DHS rate for 2019
dhs_rate_2019 <- under_5_current %>%
  filter(birth_year == 2019) %>%
  summarise(rate = mean(b5)) %>%
  pull(rate)

# Calculate missing deaths needed to reach external benchmark
missing_deaths_needed <- external_rate_2019 - dhs_rate_2019
missing_deaths_pct <- missing_deaths_needed / dhs_rate_2019

cat("\n📊 EXTERNAL BENCHMARK CALIBRATION:\n")
cat(sprintf("  External U5MR 2019 (GBD/UN): %.1f per 1000\n", external_u5mr_2019))
cat(sprintf("  DHS observed rate 2019: %.1f per 1000\n", dhs_rate_2019 * 1000))
cat(sprintf("  Missing deaths needed: %.1f per 1000\n", missing_deaths_needed * 1000))
cat(sprintf("  Required adjustment: +%.1f%%\n", missing_deaths_pct * 100))


# PART C: FIND THRESHOLD THAT ACHIEVES EXTERNAL BENCHMARK


# Focus on 2019 data for calibration
calibration_year <- 2019
calibration_data <- under_5_current %>%
  filter(birth_year == calibration_year, b5 == 0)  # Censored children only

if(nrow(calibration_data) < 10) {
  cat("\n⚠️ Insufficient data for", calibration_year, "- using nearest year\n")
  calibration_year <- 2018
  calibration_data <- under_5_current %>%
    filter(birth_year == calibration_year, b5 == 0)
}

cat(sprintf("\n📊 CALIBRATING THRESHOLD USING %d DATA:\n", calibration_year))
cat(sprintf("  Censored children in %d: %d\n", calibration_year, nrow(calibration_data)))

# Calculate how many additional deaths needed in this cohort
total_children_year <- under_5_current %>% 
  filter(birth_year == calibration_year) %>% 
  nrow()

additional_deaths_needed <- round(total_children_year * missing_deaths_needed)
additional_deaths_pct <- additional_deaths_needed / nrow(calibration_data)

cat(sprintf("  Total children in %d: %d\n", calibration_year, total_children_year))
cat(sprintf("  Additional deaths needed: %d (%.1f%% of censored)\n", 
            additional_deaths_needed, additional_deaths_pct * 100))

# Find threshold corresponding to this percentile
if(additional_deaths_pct > 0 && additional_deaths_pct < 1) {
  calibrated_threshold <- quantile(
    calibration_data$pred_risk, 
    probs = 1 - additional_deaths_pct,
    na.rm = TRUE
  )
  
  cat(sprintf("\n  ✅ CALIBRATED THRESHOLD: %.4f\n", calibrated_threshold))
  cat(sprintf("     (Top %.1f%% of censored children in %d)\n", 
              additional_deaths_pct * 100, calibration_year))
} else {
  # Fallback: use percentile-based approach
  cat("\n  ⚠️ Using percentile-based fallback (top 5% of censored)\n")
  calibrated_threshold <- quantile(
    calibration_data$pred_risk, 
    probs = 0.95,
    na.rm = TRUE
  )
  additional_deaths_pct <- 0.05
}


# PART D: VALIDATE THRESHOLD ACROSS YEARS


cat("\n📊 THRESHOLD VALIDATION ACROSS YEARS:\n")

# Test the calibrated threshold on all years
threshold_validation <- under_5_current %>%
  filter(b5 == 0) %>%  # Censored only
  group_by(birth_year) %>%
  summarise(
    n_censored = n(),
    n_above_threshold = sum(pred_risk >= calibrated_threshold, na.rm = TRUE),
    pct_above = 100 * mean(pred_risk >= calibrated_threshold, na.rm = TRUE),
    mean_risk = mean(pred_risk, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    expected_pct = additional_deaths_pct * 100,
    deviation = pct_above - expected_pct
  )

print(threshold_validation)

# Calculate stability
threshold_stability <- sd(threshold_validation$pct_above) / mean(threshold_validation$pct_above)
cat(sprintf("\n  Threshold stability (CV): %.2f\n", threshold_stability))
if(threshold_stability < 0.3) {
  cat("  ✅ Threshold relatively stable across years\n")
} else {
  cat("  ⚠️ Threshold varies substantially - consider year-specific thresholds\n")
}


# PART E: COMPARE WITH YOUDEN'S J (FOR REFERENCE ONLY)


# Calculate Youden's J for comparison (but don't use it)
roc_data <- under_5_current %>% filter(!is.na(b5))
roc_obj <- roc(roc_data$b5, roc_data$pred_risk, quiet = TRUE)
youden_threshold <- coords(roc_obj, "best", ret = "threshold", best.method = "youden")$threshold[1]
youden_pct <- 100 * mean(under_5_current$pred_risk[under_5_current$b5 == 0] >= youden_threshold, na.rm = TRUE)

cat("\n📊 COMPARISON WITH YOUDEN'S J:\n")
cat(sprintf("  Youden's J threshold: %.4f\n", youden_threshold))
cat(sprintf("  Would classify: %.1f%% of censored as deaths\n", youden_pct))
cat(sprintf("  Calibrated threshold: %.4f\n", calibrated_threshold))
cat(sprintf("  Classifies: %.1f%% of censored as deaths\n", additional_deaths_pct * 100))


# PART F: APPLY CALIBRATED THRESHOLD


# Use calibrated threshold
optimal_threshold <- calibrated_threshold

# Get performance metrics at this threshold (for information only)
perf <- coords(roc_obj, optimal_threshold, 
               ret = c("sensitivity", "specificity", "ppv", "npv", "accuracy"))

cat("\n📊 PERFORMANCE AT CALIBRATED THRESHOLD:\n")
cat(sprintf("  Sensitivity: %.1f%%\n", 100 * perf$sensitivity))
cat(sprintf("  Specificity: %.1f%%\n", 100 * perf$specificity))
cat(sprintf("  PPV: %.1f%%\n", 100 * perf$ppv))
cat(sprintf("  NPV: %.1f%%\n", 100 * perf$npv))

# Create binary prediction using CALIBRATED threshold
under_5_current$pred_death <- ifelse(under_5_current$pred_risk >= optimal_threshold, 1, 0)

# Verify adjustment in 2019
verification <- under_5_current %>%
  filter(birth_year == 2019) %>%
  summarise(
    original_rate = 1000 * mean(b5),
    adjusted_rate = 1000 * mean(ifelse(b5 == 0 & pred_death == 1, 1, b5)),
    n = n()
  )

cat("\n📊 VERIFICATION FOR 2019:\n")
cat(sprintf("  Original DHS rate: %.1f per 1000\n", verification$original_rate))
cat(sprintf("  Adjusted rate: %.1f per 1000\n", verification$adjusted_rate))
cat(sprintf("  Target external rate: %.1f per 1000\n", external_u5mr_2019))

if(abs(verification$adjusted_rate - external_u5mr_2019) < 10) {
  cat("  ✅ Calibration successful - adjusted rate matches external benchmark\n")
} else {
  cat("  ⚠️ Calibration off - consider iterative adjustment\n")
}


# PART G: SENSITIVITY ANALYSIS AROUND CALIBRATED THRESHOLD


cat("\n📊 SENSITIVITY ANALYSIS:\n")

# Test thresholds around calibrated value
sensitivity_range <- seq(
  max(0.01, calibrated_threshold - 0.03), 
  min(0.30, calibrated_threshold + 0.03), 
  length.out = 7
)

sensitivity_results <- map_df(sensitivity_range, function(t) {
  
  # Calculate 2019 rate with this threshold
  rate_2019 <- under_5_current %>%
    filter(birth_year == 2019) %>%
    mutate(test_adjusted = ifelse(b5 == 0 & pred_risk >= t, 1, b5)) %>%
    summarise(rate = 1000 * mean(test_adjusted)) %>%
    pull(rate)
  
  # Calculate overall adjusted rate
  pct_classified <- 100 * mean(under_5_current$pred_risk[under_5_current$b5 == 0] >= t, na.rm = TRUE)
  
  tibble(
    threshold = t,
    rate_2019_per_1000 = rate_2019,
    pct_censored_classified = pct_classified
  )
})

print(sensitivity_results)

# Select final threshold (use calibrated value)
cat(sprintf("\n  FINAL SELECTED THRESHOLD: %.4f\n", optimal_threshold))
cat(sprintf("  This classifies %.1f%% of censored children as deaths\n", 
            additional_deaths_pct * 100))
```

```{r step46}
# HELPER FUNCTION: Calibrate threshold to any external benchmark
calibrate_threshold_to_benchmark <- function(data, benchmark_year, benchmark_rate_per_1000) {
    benchmark_rate <- benchmark_rate_per_1000 / 1000
  
  # Get DHS rate for benchmark year
  dhs_rate <- data %>%
    filter(birth_year == benchmark_year) %>%
    summarise(rate = mean(b5)) %>%
    pull(rate)
  
  # Calculate missing deaths needed
  missing_needed <- benchmark_rate - dhs_rate
  
  # Get censored children in benchmark year
  censored_year <- data %>%
    filter(birth_year == benchmark_year, b5 == 0)
  
  total_children <- data %>% filter(birth_year == benchmark_year) %>% nrow()
  additional_deaths_needed <- round(total_children * missing_needed)
  pct_of_censored <- additional_deaths_needed / nrow(censored_year)
  
  # Find threshold
  if(pct_of_censored > 0 && pct_of_censored < 1) {
    threshold <- quantile(censored_year$pred_risk, probs = 1 - pct_of_censored, na.rm = TRUE)
  } else {
    threshold <- quantile(censored_year$pred_risk, probs = 0.95, na.rm = TRUE)
    warning("Using default 95th percentile threshold")
  }
  
  return(list(
    threshold = threshold,
    pct_classified = pct_of_censored * 100,
    dhs_rate = dhs_rate * 1000,
    target_rate = benchmark_rate_per_1000
  ))
}

# Example usage:
# result <- calibrate_threshold_to_benchmark(under_5_current, 2019, 55)
# optimal_threshold <- result$threshold
```

```{r step47}
# STEP 7: IDENTIFY PREDICTED DEATHS AMONG CENSORED


cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 7: IDENTIFYING PREDICTED DEATHS AMONG CENSORED\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

censored_children <- under_5_current %>% filter(b5 == 0)
predicted_deaths <- censored_children %>% filter(pred_death == 1)

n_predicted <- nrow(predicted_deaths)
pct_censored <- 100 * n_predicted / nrow(censored_children)

cat("\n📊 PREDICTED DEATHS AMONG CENSORED:\n")
cat(sprintf("  Censored children: %d\n", nrow(censored_children)))
cat(sprintf("  Predicted to die by age 5: %d (%.1f%% of censored)\n", 
            n_predicted, pct_censored))

# By cohort
pred_by_cohort <- censored_children %>%
  group_by(birth_cohort) %>%
  summarise(
    Censored_N = n(),
    Predicted_Deaths = sum(pred_death == 1),
    Percent_Predicted = 100 * mean(pred_death == 1),
    Mean_Risk = mean(pred_risk),
    .groups = 'drop'
  )

cat("\n📊 PREDICTED DEATHS BY COHORT:\n")
print(pred_by_cohort)


```

```{r step48}
# STEP 8: CREATE ADJUSTED MORTALITY VARIABLE WITH PREDICTIONS
# (CORRECTED VERSION - No row multiplication)

cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 8: CREATING ADJUSTED MORTALITY ESTIMATES\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# Start with original filtered dataset (89,069 children, 2000-2019 only)
final_data <- cens_2000_2019

cat("\n📊 BEFORE MERGE:\n")
cat(sprintf("  Final data rows: %d\n", nrow(final_data)))
cat(sprintf("  Under-5 predictions: %d\n", nrow(under_5_current)))

# FIX: Add DISTINCT to remove duplicates in prediction data
predictions_unique <- under_5_current %>%
  # Create a unique identifier if not already present
  mutate(unique_id = paste(b7, b19, birth_year, Clu_id, sep = "_")) %>%
  # Keep only unique combinations of the merge keys
  distinct(b7, b19, birth_year, Clu_id, .keep_all = TRUE) %>%
  select(b7, b19, birth_year, Clu_id, pred_survival, pred_risk, pred_death, unique_id)

cat(sprintf("  Unique predictions after deduplication: %d\n", nrow(predictions_unique)))

# Check for duplicates before merging
duplicates_check <- predictions_unique %>%
  group_by(b7, b19, birth_year, Clu_id) %>%
  summarise(n = n(), .groups = 'drop') %>%
  filter(n > 1)

if(nrow(duplicates_check) > 0) {
  cat("⚠️ Warning: Still found", nrow(duplicates_check), "duplicate key combinations\n")
} else {
  cat("✅ No duplicate keys - safe to merge\n")
}

# Merge predictions (NOW using distinct data)
final_data <- final_data %>%
  left_join(
    predictions_unique %>%
      select(b7, b19, birth_year, Clu_id, pred_survival, pred_risk, pred_death),
    by = c("b7", "b19", "birth_year", "Clu_id"),
    relationship = "many-to-one"  # Explicitly state expected relationship
  )

# VALIDATION: Check that rows didn't multiply
cat("\n📊 AFTER MERGE VALIDATION:\n")
cat(sprintf("  Final data rows: %d\n", nrow(final_data)))
cat(sprintf("  Original data rows: %d\n", nrow(cens_2000_2019)))

if(nrow(final_data) == nrow(cens_2000_2019)) {
  cat("✅ SUCCESS: Row count preserved (no multiplication)\n")
} else {
  cat("❌ ERROR: Row count changed from", nrow(cens_2000_2019), 
      "to", nrow(final_data), "\n")
  stop("Merge failed - check for duplicates in key variables")
}

# Check how many predictions were successfully merged
predictions_merged <- sum(!is.na(final_data$pred_survival))
cat(sprintf("  Predictions merged: %d of %d under-5 children\n", 
            predictions_merged, nrow(under_5_current)))

# Create adjusted mortality variable
final_data$b5_adjusted <- final_data$b5_original
final_data$b5_adjusted[final_data$b5_original == 0 & 
                       final_data$pred_death == 1] <- 1

# VALIDATION 8.1: Overall adjustment impact
orig_deaths <- sum(final_data$b5_original == 1, na.rm = TRUE)
adj_deaths <- sum(final_data$b5_adjusted == 1, na.rm = TRUE)
orig_rate <- 100 * mean(final_data$b5_original == 1, na.rm = TRUE)
adj_rate <- 100 * mean(final_data$b5_adjusted == 1, na.rm = TRUE)

cat("\n📊 OVERALL ADJUSTMENT IMPACT:\n")
cat(sprintf("  Original deaths: %d (%.1f%%)\n", orig_deaths, orig_rate))
cat(sprintf("  Adjusted deaths: %d (%.1f%%)\n", adj_deaths, adj_rate))
cat(sprintf("  Additional deaths identified: %d\n", adj_deaths - orig_deaths))
cat(sprintf("  Relative increase: %.1f%%\n", 
            100 * (adj_rate - orig_rate) / orig_rate))

# VALIDATION 8.2: Adjustment by cohort
adjustment_by_cohort <- final_data %>%
  group_by(birth_cohort) %>%
  summarise(
    N = n(),
    Original_Deaths = sum(b5_original == 1, na.rm = TRUE),
    Adjusted_Deaths = sum(b5_adjusted == 1, na.rm = TRUE),
    Original_Rate = 100 * mean(b5_original == 1, na.rm = TRUE),
    Adjusted_Rate = 100 * mean(b5_adjusted == 1, na.rm = TRUE),
    Censoring_Rate = 100 * mean(b19 < 60 & b5_original == 0, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    Rate_Change = Adjusted_Rate - Original_Rate,
    Pct_Increase = 100 * (Adjusted_Rate - Original_Rate) / Original_Rate
  )

cat("\n📊 ADJUSTMENT BY BIRTH COHORT:\n")
print(adjustment_by_cohort)

# Save the corrected dataset
saveRDS(final_data, paste0(CENSOR_CENSOR_DIR, "final_data_corrected.rds"), compress = TRUE)
if(!FAST_MODE) {
  write.csv(final_data, paste0(CENSOR_CENSOR_DIR, "final_data_corrected.csv"), row.names = FALSE)
}

cat("\n✅ Corrected final data saved with", nrow(final_data), "rows (2000-2019 only)\n")
```

```{r step49}
# STEP 10: SAVE FINAL DATASETS
cat("\n", paste(rep("═", 70), collapse = ""), "\n")
cat("STEP 10: SAVING FINAL DATASETS\n")
cat(paste(rep("═", 70), collapse = ""), "\n")

# Save RDS (compressed binary)
saveRDS(final_data, 
        paste0(CENSOR_CENSOR_DIR, "final_data_optimized.rds"),
        compress = TRUE)

if(!FAST_MODE) {
  write.csv(final_data, 
            paste0(CENSOR_CENSOR_DIR, "final_data_optimized.csv"), 
            row.names = FALSE)
}

cat("\n✅ Files saved:\n")
cat("  - final_data_optimized.rds (R binary format)\n")
if(!FAST_MODE) cat("  - final_data_optimized.csv (CSV format)\n")
cat(sprintf("\n  Final dataset: %d children\n", nrow(final_data)))
```

```{r step50}
# Calculate death counts and percentages
years <- 2000:2019
original_deaths <- c(758,814,759,775,708,588,487,531,465,422,437,260,241,219,193,163,87,69,59,13)
adjusted_deaths <- c(893,1024,940,935,974,627,645,708,617,570,514,361,367,339,340,265,147,103,78,17)
difference <- adjusted_deaths - original_deaths
pct_increase <- round((difference/original_deaths)*100, 1)

# Create summary dataframe
comparison <- data.frame(
  year = years,
  original = original_deaths,
  adjusted = adjusted_deaths,
  diff = difference,
  pct_increase = pct_increase
)
```

```{r session-info}
# Save session information for reproducibility
sink(paste0(CENSOR_CENSOR_DIR, "session_info.txt"))
cat("========================================================\n")
cat("SESSION INFORMATION\n")
cat("========================================================\n\n")
sessionInfo()
sink()

cat("\n✅ Session info saved to:", paste0(CENSOR_CENSOR_DIR, "session_info.txt"), "\n")
cat("\n========================================================\n")
cat("🎯 ANALYSIS COMPLETE - READY FOR BAYESIAN MODELING\n")
cat("========================================================\n")
```


