---
title: "Untitled"
output: html_document
date: "2026-01-06"
---

# Covariate Preparation Pipeline for 5 km Under‑Five Mortality Mapping in (2000‑2015)

IMPORTANT – RAW DATA NOT INCLUDED, BUT PROCESSED OUTPUTS FOR ALL YEARS (2000‑2015) ARE PROVIDED

This script processes freely available raw rasters (ERA5, GBD, MODIS, WorldPop, Malaria Atlas Project, etc.). These raw data are NOT included in this deposit due to their large size. The folder paths (e.g., D:/new resea/Ethiopia/...) reflect the author’s local machine; users must download the raw data themselves and adjust the paths if they wish to re‑run the script.

However, the final outputs of this script – adjusted covariates (21 variables reduced to 16 after correlation filtering), 5 km population rasters, and inhabited area masks for ALL years 2000‑2015 – ARE included in this deposit under the `adjusted_covariates/` folder. These ready‑to‑use GeoTIFFs can be directly employed in downstream analyses (e.g., censoring adjustment or INLA modelling) without re‑running the script.

The script is shared for methodological transparency and to document every processing step (log1p, z‑score, capping, metadata generation). Users who wish to reproduce the entire workflow or process additional years can do so after obtaining the raw data and modifying the file paths.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Covariate Preparation Pipeline for 5 km Under‑Five Mortality Mapping in Ethiopia (2000‑2015)

# Dignosis for transformation decation 

```{r}
folder_path <- "D:/new resea/Ethiopia/ERA5 Ethiopia/dewpoi_temp_2m"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (dewPoi_00.tif to dewPoi_19.tif)
files <- sprintf("dewPoi_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_dewpoint <- rast(rasters)
# Pool all values into a single vector
values_dewpoint <- values(stack_dewpoint, mat = FALSE)
values_dewpoint <- values_dewpoint[!is.na(values_dewpoint)]
# Compute diagnostics
mean_val <- mean(values_dewpoint)
median_val <- median(values_dewpoint)
min_val <- min(values_dewpoint)
max_val <- max(values_dewpoint)
skew_val <- skewness(values_dewpoint)
percent_zeros <- sum(values_dewpoint == 0)/length(values_dewpoint) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "dewpoint_temp_2m",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "DewPoint_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```


```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/deareha"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (2000–2017 -> dear_et_00.tif to dear_et_17.tif)
files <- sprintf("dear_et_%02d.tif", 0:17)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_deareha <- rast(rasters)
# Pool all values into a single vector
values_deareha <- values(stack_deareha, mat = FALSE)
values_deareha <- values_deareha[!is.na(values_deareha)]
# Compute diagnostics
mean_val <- mean(values_deareha)
median_val <- median(values_deareha)
min_val <- min(values_deareha)
max_val <- max(values_deareha)
skew_val <- skewness(values_deareha)
percent_zeros <- sum(values_deareha == 0)/length(values_deareha) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "deareha",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Dear_Diagnosis_2000_2017.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/ERA5 Ethiopia/solarradationdown"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (rad_00.tif to rad_19.tif)
files <- sprintf("rad_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_solar <- rast(rasters)
# Pool all values into a single vector
values_solar <- values(stack_solar, mat = FALSE)
values_solar <- values_solar[!is.na(values_solar)]
# Compute diagnostics
mean_val <- mean(values_solar)
median_val <- median(values_solar)
min_val <- min(values_solar)
max_val <- max(values_solar)
skew_val <- skewness(values_solar)
percent_zeros <- sum(values_solar == 0)/length(values_solar) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "solarradationdown",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "SolarRad_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/ERA5 Ethiopia/potentialevaporation"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (evap_00.tif to evap_19.tif)
files <- sprintf("evap_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_evap <- rast(rasters)
# Pool all values into a single vector
values_evap <- values(stack_evap, mat = FALSE)
values_evap <- values_evap[!is.na(values_evap)]
# Compute diagnostics
mean_val <- mean(values_evap)
median_val <- median(values_evap)
min_val <- min(values_evap)
max_val <- max(values_evap)
skew_val <- skewness(values_evap)
percent_zeros <- sum(values_evap == 0)/length(values_evap) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "potentialevaporation",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Evaporation_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/ERA5 Ethiopia/runoff"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (run_00.tif to run_19.tif)
files <- sprintf("run_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_runoff <- rast(rasters)
# Pool all values into a single vector
values_runoff <- values(stack_runoff, mat = FALSE)
values_runoff <- values_runoff[!is.na(values_runoff)]
# Compute diagnostics
mean_val <- mean(values_runoff)
median_val <- median(values_runoff)
min_val <- min(values_runoff)
max_val <- max(values_runoff)
skew_val <- skewness(values_runoff)
percent_zeros <- sum(values_runoff == 0)/length(values_runoff) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "runoff",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Runoff_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/NDVI Ethiopia"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (NDVI_2000_Ethiopia.tif to NDVI_2019_Ethiopia.tif)
years <- 2000:2019
files <- sprintf("NDVI_%d_Ethiopia.tif", years)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_ndvi <- rast(rasters)
# Pool all values into a single vector
values_ndvi <- values(stack_ndvi, mat = FALSE)
values_ndvi <- values_ndvi[!is.na(values_ndvi)]
# Compute diagnostics
mean_val <- mean(values_ndvi)
median_val <- median(values_ndvi)
min_val <- min(values_ndvi)
max_val <- max(values_ndvi)
skew_val <- skewness(values_ndvi)
percent_zeros <- sum(values_ndvi == 0)/length(values_ndvi) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "NDVI",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "NDVI_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/pm2.5"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (pm25_et_00.tif to pm25_et_19.tif)
files <- sprintf("pm25_et_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_pm25 <- rast(rasters)
# Pool all values into a single vector
values_pm25 <- values(stack_pm25, mat = FALSE)
values_pm25 <- values_pm25[!is.na(values_pm25)]
# Compute diagnostics
mean_val <- mean(values_pm25)
median_val <- median(values_pm25)
min_val <- min(values_pm25)
max_val <- max(values_pm25)
skew_val <- skewness(values_pm25)
percent_zeros <- sum(values_pm25 == 0)/length(values_pm25) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "PM2.5",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "PM25_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/malaria"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (malar_et_00.tif to malar_et_18.tif)
files <- sprintf("malar_et_%02d.tif", 0:18)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_malaria <- rast(rasters)
# Pool all values into a single vector
values_malaria <- values(stack_malaria, mat = FALSE)
values_malaria <- values_malaria[!is.na(values_malaria)]
# Compute diagnostics
mean_val <- mean(values_malaria)
median_val <- median(values_malaria)
min_val <- min(values_malaria)
max_val <- max(values_malaria)
skew_val <- skewness(values_malaria)
percent_zeros <- sum(values_malaria == 0)/length(values_malaria) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "malaria",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Malaria_Diagnosis_2000_2018.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
folder_path <- "D:/new resea/Ethiopia/deareha"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (dear_et_00.tif to dear_et_17.tif)
files <- sprintf("dear_et_%02d.tif", 0:17)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_deareha <- rast(rasters)
# Pool all values into a single vector
values_deareha <- values(stack_deareha, mat = FALSE)
values_deareha <- values_deareha[!is.na(values_deareha)]
# Compute diagnostics
mean_val <- mean(values_deareha)
median_val <- median(values_deareha)
min_val <- min(values_deareha)
max_val <- max(values_deareha)
skew_val <- skewness(values_deareha)
percent_zeros <- sum(values_deareha == 0)/length(values_deareha) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "deareha",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "Dear_Diagnosis_2000_2017.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# arameters
folder_path <- "D:/new resea/Ethiopia/HIV"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (HIV_ETH_00.tif to HIV_ETH_19.tif)
files <- sprintf("HIV_ETH_%02d.tif", 0:19)
file_paths <- file.path(folder_path, files)
# Keep only existing files
file_paths <- file_paths[file.exists(file_paths)]
# Load first layer of each raster safely
rasters <- lapply(file_paths, function(f){
  r <- rast(f, lyrs = 1)
  return(r)
})
# Stack rasters
stack_hiv <- rast(rasters)
# Pool all values into a single vector
values_hiv <- values(stack_hiv, mat = FALSE)
values_hiv <- values_hiv[!is.na(values_hiv)]
# Compute diagnostics
mean_val <- mean(values_hiv)
median_val <- median(values_hiv)
min_val <- min(values_hiv)
max_val <- max(values_hiv)
skew_val <- skewness(values_hiv)
percent_zeros <- sum(values_hiv == 0)/length(values_hiv) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "HIV",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "HIV_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/Sever wasting"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (SW 2000.tif to SW 2019.tif)
years <- 2000:2019
files <- sprintf("SW %d.tif", years)   # note the space instead of underscore
file_paths <- file.path(folder_path, files)
# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_sw <- rast(rasters)
# Pool all values into a single vector
values_sw <- values(stack_sw, mat = FALSE)
values_sw <- values_sw[!is.na(values_sw)]
# Compute diagnostics
mean_val <- mean(values_sw)
median_val <- median(values_sw)
min_val <- min(values_sw)
max_val <- max(values_sw)
skew_val <- skewness(values_sw)
percent_zeros <- sum(values_sw == 0)/length(values_sw) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "Severe Wasting",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "SevereWasting_Diagnosis_2000_2019.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters
folder_path <- "D:/new resea/Ethiopia/dpt3"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (DPT3_ET_00.tif to DPT3_ET_16.tif)
files <- sprintf("DPT3_ET_%02d.tif", 0:16)
file_paths <- file.path(folder_path, files)

# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}
# Load first layer safely
rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})
# Stack rasters
stack_dpt3 <- rast(rasters)
# Pool all values into a single vector
values_dpt3 <- values(stack_dpt3, mat = FALSE)
values_dpt3 <- values_dpt3[!is.na(values_dpt3)]
# Compute diagnostics
mean_val <- mean(values_dpt3)
median_val <- median(values_dpt3)
min_val <- min(values_dpt3)
max_val <- max(values_dpt3)
skew_val <- skewness(values_dpt3)
percent_zeros <- sum(values_dpt3 == 0)/length(values_dpt3) * 100
# Decide transformation based on pooled skewness
if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}
# Create diagnosis table
diagnosis_table <- data.frame(
  Covariate = "DPT3",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)
# Save table
output_file <- file.path(output_path, "DPT3_Diagnosis_2000_2016.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
# Parameters

folder_path <- "D:/new resea/Ethiopia/wash"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)


# Generate file paths explicitly (wash 2000.tif to wash 2017.tif)

years <- 2000:2017
files <- sprintf("wash %d.tif", years)   # note the space
file_paths <- file.path(folder_path, files)

# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}


# Load first layer safely

rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})

# Stack rasters
stack_wash <- rast(rasters)


# Pool all values into a single vector

values_wash <- values(stack_wash, mat = FALSE)
values_wash <- values_wash[!is.na(values_wash)]


# Compute diagnostics

mean_val <- mean(values_wash)
median_val <- median(values_wash)
min_val <- min(values_wash)
max_val <- max(values_wash)
skew_val <- skewness(values_wash)
percent_zeros <- sum(values_wash == 0)/length(values_wash) * 100


# Decide transformation based on pooled skewness

if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}


# Create diagnosis table

diagnosis_table <- data.frame(
  Covariate = "WASH",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)


# Save table

output_file <- file.path(output_path, "WASH_Diagnosis_2000_2017.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
library(terra)
library(moments)
# Parameters
folder_path <- "D:/new resea/Ethiopia/water"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)
# Generate file paths explicitly (water 2000.tif to water 2017.tif)
years <- 2000:2017
files <- sprintf("water %d.tif", years)   # note the space
file_paths <- file.path(folder_path, files)

# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}


# Load first layer safely

rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})

# Stack rasters
stack_water <- rast(rasters)


# Pool all values into a single vector

values_water <- values(stack_water, mat = FALSE)
values_water <- values_water[!is.na(values_water)]


# Compute diagnostics

mean_val <- mean(values_water)
median_val <- median(values_water)
min_val <- min(values_water)
max_val <- max(values_water)
skew_val <- skewness(values_water)
percent_zeros <- sum(values_water == 0)/length(values_water) * 100


# Decide transformation based on pooled skewness

if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}


# Create diagnosis table

diagnosis_table <- data.frame(
  Covariate = "Water",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)


# Save table

output_file <- file.path(output_path, "Water_Diagnosis_2000_2017.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```

```{r}
library(terra)
library(moments)


# Parameters

folder_path <- "D:/new resea/Ethiopia/education"
output_path <- "D:/new resea/Ethiopia/data managment/diagnose"
dir.create(output_path, recursive = TRUE, showWarnings = FALSE)


# Generate file paths explicitly (Educ_2000.tif to Educ_2017.tif)

years <- 2000:2017
files <- sprintf("Educ_%d.tif", years)
file_paths <- file.path(folder_path, files)

# Keep only existing files
existing_files <- file_paths[file.exists(file_paths)]
if(length(existing_files) == 0){
  stop("No raster files found in folder: ", folder_path)
}


# Load first layer safely

rasters <- lapply(existing_files, function(f){
  rast(f, lyrs = 1)
})

# Stack rasters
stack_educ <- rast(rasters)


# Pool all values into a single vector

values_educ <- values(stack_educ, mat = FALSE)
values_educ <- values_educ[!is.na(values_educ)]


# Compute diagnostics

mean_val <- mean(values_educ)
median_val <- median(values_educ)
min_val <- min(values_educ)
max_val <- max(values_educ)
skew_val <- skewness(values_educ)
percent_zeros <- sum(values_educ == 0)/length(values_educ) * 100


# Decide transformation based on pooled skewness

if(abs(skew_val) < 1){
  rec <- "No transform"
} else if(abs(skew_val) < 2){
  rec <- "Usually no transform, inspect tail"
} else if(abs(skew_val) < 3){
  rec <- "Consider log1p or sqrt transform"
} else {
  rec <- "Strongly consider transform (log1p)"
}


# Create diagnosis table

diagnosis_table <- data.frame(
  Covariate = "Education",
  Global_Mean = mean_val,
  Global_Median = median_val,
  Min = min_val,
  Max = max_val,
  Skewness = skew_val,
  Percent_Zeros = percent_zeros,
  Recommendation = rec,
  stringsAsFactors = FALSE
)


# Save table

output_file <- file.path(output_path, "Education_Diagnosis_2000_2017.csv")
write.csv(diagnosis_table, output_file, row.names = FALSE)
cat("Diagnosis table saved to:", output_file, "\n")
```


####################################################################################################################

# COVARIATE PREPARATION PIPELINE FOR UNDER-FIVE MORTALITY PREDICTION

### Night time lights (NTL – Binary Urban/Rural

```{r}
cat("\n=== Night-Time Lights 2000 PROCESSING (Binary Urban/Rural for INLA + SPDE) ===\n")

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD ETHIOPIA BOUNDARY (for masking)
ethiopia_poly <- vect("D:/shap/ethiopia/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Ethiopia boundary loaded and projected\n")

# 3. LOAD RAW NIGHT-TIME LIGHT RASTER (2000)
ntl_raw <- rast("D:/new resea/Ethiopia/nightlight/light 2000.tif")
if(nlyr(ntl_raw) > 1) ntl_raw <- ntl_raw[[1]]
cat("Raw NTL raster loaded\n")
cat("Value range:", range(values(ntl_raw), na.rm=TRUE), "\n\n")

# 4. PROJECT TO TEMPLATE CRS
ntl_proj <- if(!same.crs(ntl_raw, template)) project(ntl_raw, template) else ntl_raw
cat("Projection complete\n")

# 5. RESAMPLE TO 5 KM TEMPLATE
ntl_5km <- resample(ntl_proj, template, method="bilinear")
cat("Resampled to 5km resolution\n")

# 6. CROP + MASK TO ETHIOPIA
ntl_crop <- crop(ntl_5km, ethiopia_poly_utm)
ntl_clip <- mask(ntl_crop, ethiopia_poly_utm)
cat("Clipped to Ethiopia boundary\n")

# 7. CREATE BINARY URBAN VARIABLE
# DN > 0 = 1 (urban)
# DN = 0  = 0 (rural)
ntl_binary <- classify(
  ntl_clip,
  rcl = matrix(c(-Inf, 0, 0,
                 0, Inf, 1),
               ncol=3, byrow=TRUE)
)
names(ntl_binary) <- "NTL_Urban_Binary"
cat("Binary classification complete (0 = rural, 1 = urban)\n")

# 8. FINAL CHECK
ntl_vals <- values(ntl_binary)
ntl_vals_nonNA <- ntl_vals[!is.na(ntl_vals)]

cat("Valid pixels:", length(ntl_vals_nonNA), "\n")
cat("Urban proportion:", mean(ntl_vals_nonNA), "\n\n")

# 9. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/NTL00_5km_binary.tif"

writeRaster(
  ntl_binary,
  filename = output_raster,
  overwrite = TRUE,
  gdal = "COMPRESS=LZW"
)

cat("Binary NTL raster saved:", output_raster, "\n")

# 10. METADATA
metadata <- list(
  variable_name = "Night-Time Lights (Binary Urban/Rural)",
  variable_code = "NTL00_5km_binary",
  year = 2000,
  country = "Ethiopia",
  
  source = "DMSP-OLS Night-Time Lights",
  original_units = "Digital Number (DN)",
  
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  
  crs = as.character(crs(ntl_binary)),
  resolution_m = paste(res(ntl_binary), collapse=" x "),
  n_cells_total = ncell(ntl_binary),
  n_cells_valid = length(ntl_vals_nonNA),
  n_cells_missing = sum(is.na(ntl_vals)),
  
  classification_rule = "DN > 0 = 1 (urban), DN = 0 = 0 (rural)",
  proportion_urban = mean(ntl_vals_nonNA),
  proportion_rural = 1 - mean(ntl_vals_nonNA),
  unique_values = sort(unique(ntl_vals_nonNA)),
  
  transformation = "none (binary classification)",
  standardization = "none",
  imputation = "none",
  
  inla_ready = TRUE,
  spde_alignment_checked = TRUE
)

jsonlite::write_json(
  metadata,
  path = gsub(".tif$","_metadata.json",output_raster),
  pretty = TRUE,
  auto_unbox = TRUE
)

cat("Metadata saved\n")

cat("=== Night-Time Lights 2000 PROCESSING COMPLETE ===\n")
cat("Raster:", output_raster, "\n")
cat("Metadata JSON:", gsub(".tif$","_metadata.json",output_raster), "\n")
```


###  Dewpoint Temperature at 2 meters above ground level (K)  ERA5 Unit: Kelvin (K)

DEWPOINT TEMPERATURE COVARIATE PREPARATION
For Bayesian Hierarchical Model with INLA + SPDE
Target: Under-five mortality prediction at 5km resolution, Ethiopia 2000
  Source: ERA5 Reanalysis Original units: Kelvin (K)
Processing date: 2026-02-26

```{r}
cat("\n=== DEWPOINT (2m) PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

# 1. LOAD TEMPLATE AND STUDY AREA
library(terra)
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
ethiopia_poly <- vect("D:/shap/ethiopia/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW ERA5 DEWPOINT (Kelvin)
dew_raw <- rast("D:/new resea/Ethiopia/ERA5 Ethiopia/dewpoi_temp_2m/dewPoi_00.tif")
if (nlyr(dew_raw) > 1) dew_raw <- dew_raw[[1]]
cat("Raw dewpoint loaded\n")

# 3. PROJECT + RESAMPLE TO 5km GRID
dew_proj <- if (!same.crs(dew_raw, template)) project(dew_raw, template) else dew_raw
dew_5km <- resample(dew_proj, template, method = "bilinear")
cat("Projected and resampled to 5km\n")

# 4. CLIP TO ETHIOPIA
dew_crop <- crop(dew_5km, ethiopia_poly_utm)
dew_clip <- mask(dew_crop, ethiopia_poly_utm)
cat("Clipped to Ethiopia boundary\n")

# 5. Z-SCORE STANDARDIZATION (Ethiopia pixels only)
vals <- values(dew_clip)
vals_eth <- vals[!is.na(vals)]
vals_z <- (vals - mean(vals_eth)) / sd(vals_eth)
values(dew_clip) <- vals_z
cat("Standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 6. GEOMETRY CHECK
geom_check <- compareGeom(dew_clip, template, stopOnError = FALSE)
if (!geom_check) stop("Geometry mismatch with template.")
cat("Geometry alignment: PASS\n")

# 7. SAVE COVARIATE RASTER
output_filename <- "eth_2000_dewpoint_5km_z.tif"
output_path <- file.path("D:/new resea/Ethiopia/each year/2000/Adjusted_Raster", output_filename)
writeRaster(dew_clip, filename = output_path, overwrite = TRUE, gdal = "COMPRESS=LZW")
cat("Raster saved:", output_path, "\n")

# 8. METADATA FOR COVARIATE STAGE ONLY
metadata <- list(
  variable_name = "Dewpoint Temperature at 2m",
  variable_code = "dewpoint",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5 Reanalysis",
  source_doi = "10.24381/cds.adbb2d47",
  original_units = "Kelvin",
  covariate_processing = list(
    projection = "EPSG:32637 (UTM Zone 37N)",
    resampling = "Bilinear",
    masking = "Masked to Ethiopia national boundary",
    missing_handling = "NA values retained; Ethiopia pixels used for standardization",
    scaling = "Z-score standardization (mean=0, SD=1) based on Ethiopia pixels only"
  ),
  spatial_resolution_m = paste(round(res(dew_clip), 2), collapse = " x "),
  crs = as.character(crs(dew_clip)),
  n_cells = ncell(dew_clip),
  processing_steps = "projected, resampled (bilinear), masked to Ethiopia, standardized (z-score)",
  alignment_passed = geom_check,
  processing_date = as.character(Sys.Date())
)

# Save metadata
jsonlite::write_json(
  metadata, 
  path = gsub(".tif$", "_metadata.json", output_path),
  pretty = TRUE,
  auto_unbox = TRUE
)
cat("Covariate-stage metadata saved\n")
cat("\n=== COVARIATE PREPARATION COMPLETE ===\n")
```


# malaria prevalence / risk

```{r}
cat("\n=== MALARIA PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

library(terra)
library(e1071)

# 1. LOAD TEMPLATE AND STUDY AREA
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
ethiopia_poly <- vect("D:/shap/ethiopia/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW MALARIA DATA
mal_raw <- rast("D:/new resea/Ethiopia/malaria/malar_et_00.tif")
if (nlyr(mal_raw) > 1) mal_raw <- mal_raw[[1]]
cat("Raw malaria raster loaded\n")

# 3. PROJECT + RESAMPLE TO 5km GRID
mal_proj <- if (!same.crs(mal_raw, template)) project(mal_raw, template) else mal_raw
mal_5km <- resample(mal_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 4. CLIP TO ETHIOPIA
mal_crop <- crop(mal_5km, ethiopia_poly_utm)
mal_clip <- mask(mal_crop, ethiopia_poly_utm)
cat("Clipped to Ethiopia boundary\n")

# 5. LOG1P TRANSFORMATION
vals <- values(mal_clip)
vals_eth <- vals[!is.na(vals)]
vals_log <- log1p(vals_eth)
cat("Log1p transformation applied\n")

# 6. Z-SCORE STANDARDIZATION
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std <- (vals_log - mean_val) / sd_val

# 6a. HANDLE EXTREME Z-SCORES
z_cap <- 5
n_extreme <- sum(abs(vals_std) > z_cap)
vals_std[vals_std > z_cap] <- z_cap
vals_std[vals_std < -z_cap] <- -z_cap
cat("Extreme z-scores capped at ±", z_cap, " (", n_extreme, "pixels adjusted)\n", sep="")

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(mal_clip) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1, extremes capped)\n")

# 7. GEOMETRY CHECK
geom_check <- compareGeom(mal_clip, template, stopOnError = FALSE)
if (!geom_check) stop("Geometry mismatch with template.")
cat("Geometry alignment: PASS\n")
# 8. SAVE COVARIATE RASTER
output_filename <- "eth_2000_malaria_5km_zlog_capped.tif"
output_path <- file.path("D:/new resea/Ethiopia/each year/2000/Adjusted_Raster", output_filename)
writeRaster(mal_clip, filename = output_path, overwrite = TRUE, gdal = "COMPRESS=LZW")
cat("Raster saved:", output_path, "\n")
# 9. METADATA FOR COVARIATE STAGE ONLY
metadata <- list(
  variable_name = "Malaria",
  variable_code = "malaria",
  year = 2000,
  country = "Ethiopia",
  source = "Malaria Atlas Project",
  original_units = "continuous",
  covariate_processing = list(
    projection = "EPSG:32637 (UTM Zone 37N)",
    resampling = "Bilinear",
    masking = "Masked to Ethiopia national boundary",
    transformation = "log1p",
    scaling = "Z-score standardization (mean=0, SD=1, extremes capped at ±5)"
  ),
  spatial_resolution_m = paste(round(res(mal_clip), 2), collapse = " x "),
  crs = as.character(crs(mal_clip)),
  n_cells = ncell(mal_clip),
  n_cells_valid = sum(!is.na(values(mal_clip))),
  processing_steps = "projected, resampled (bilinear), masked to Ethiopia, log1p transformed, standardized (z-score), extreme z-scores capped at ±5",
  alignment_passed = geom_check,
  processing_date = as.character(Sys.Date())
)

jsonlite::write_json(
  metadata,
  path = gsub(".tif$", "_metadata.json", output_path),
  pretty = TRUE, auto_unbox = TRUE
)
cat("Covariate-stage metadata saved\n")
cat("\n=== COVARIATE PREPARATION COMPLETE ===\n")
```




# surface solar radiation downwards (likely in J/m² over a time period, based on ERA5)

```{r}
cat("\n=== SOLAR RADIATION PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

library(terra)
library(jsonlite)

# 1. LOAD TEMPLATE AND STUDY AREA
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
ethiopia_poly <- vect("D:/shap/ethiopia/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW SOLAR RADIATION
solar_raw <- rast("D:/new resea/Ethiopia/ERA5 Ethiopia/solarradationdown/rad_00.tif")
if(nlyr(solar_raw) > 1) solar_raw <- solar_raw[[1]]
cat("Raw solar raster loaded\n")

# 3. PROJECT + RESAMPLE TO 5KM GRID
solar_proj <- if(!same.crs(solar_raw, template)) project(solar_raw, template) else solar_raw
solar_5km <- resample(solar_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 4. CROP + MASK TO ETHIOPIA
solar_crop <- crop(solar_5km, ethiopia_poly_utm)
solar_clip <- mask(solar_crop, ethiopia_poly_utm)
vals_before <- values(solar_clip)
vals_eth <- vals_before[!is.na(vals_before)]
cat("Clipped to Ethiopia, valid pixels:", length(vals_eth), "\n")

# 5. Z-SCORE STANDARDIZATION
mean_val <- mean(vals_eth)
sd_val <- sd(vals_eth)
vals_std_eth <- (vals_eth - mean_val) / sd_val

vals_std_full <- rep(NA, length(vals_before))
vals_std_full[!is.na(vals_before)] <- vals_std_eth
values(solar_clip) <- vals_std_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 6. GEOMETRY CHECK
geom_check <- compareGeom(solar_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 7. SAVE FINAL RASTER
output_filename <- "eth_2000_solar_5km_zscore.tif"
output_path <- file.path("D:/new resea/Ethiopia/each year/2000/Adjusted_Raster", output_filename)
writeRaster(solar_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_path, "\n")

# 8. METADATA
metadata <- list(
  variable_name = "Surface Solar Radiation Downwards",
  variable_code = "solar",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5",
  source_doi = "10.24381/cds.adbb2d47",
  original_units = "J/m²",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(solar_clip)),
  resolution_m = paste(res(solar_clip), collapse=" x "),
  n_cells = ncell(solar_clip),
  n_cells_valid = length(vals_eth),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_eth),
  final_max = max(vals_eth),
  final_mean = mean(vals_eth),
  final_sd = sd(vals_eth),
  final_skewness = e1071::skewness(vals_eth),
  final_kurtosis = e1071::kurtosis(vals_eth),
  alignment_passed = geom_check,
  missing_final = sum(is.na(vals_before))
)

jsonlite::write_json(metadata, path=gsub(".tif$","_metadata.json",output_path), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
cat("\n=== COVARIATE PREPARATION COMPLETE ===\n")
```


#  severe wasting prevalence at a certain spatial resolution

```{r}
cat("\n=== SEVERE WASTING (2000) PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

library(terra)
library(jsonlite)
library(e1071)

# 1. LOAD TEMPLATE AND STUDY AREA
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
ethiopia_poly <- vect("D:/shap/ethiopia/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW SEVERE WASTING RASTER
SW_raw <- rast("D:/new resea/Ethiopia/Sever wasting/SW 2000.tif")
if(nlyr(SW_raw) > 1) SW_raw <- SW_raw[[1]]
cat("Raw Severe Wasting raster loaded\n")

# 3. LOG1P TRANSFORMATION
vals <- values(SW_raw)
vals_eth <- vals[!is.na(vals)]
vals_log <- log1p(vals_eth)
cat("Log1p transformation applied\n")

# 4. Z-SCORE STANDARDIZATION
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std <- (vals_log - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(SW_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 5. PROJECT + RESAMPLE TO 5KM TEMPLATE
SW_proj <- if(!same.crs(SW_raw, template)) project(SW_raw, template) else SW_raw
SW_5km <- resample(SW_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 6. CROP + MASK TO ETHIOPIA
SW_crop <- crop(SW_5km, ethiopia_poly_utm)
SW_clip <- mask(SW_crop, ethiopia_poly_utm)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(SW_clip))), "\n")

# 7. GEOMETRY CHECK
geom_check <- compareGeom(SW_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 8. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/SW00_5km_zlog.tif"
writeRaster(SW_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 9. METADATA
vals_final <- values(SW_clip)[!is.na(values(SW_clip))]
metadata <- list(
  variable_name = "Severe Wasting Prevalence",
  variable_code = "SW00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(SW_clip)),
  resolution_m = paste(res(SW_clip), collapse=" x "),
  n_cells = ncell(SW_clip),
  n_cells_valid = length(vals_final),
  transformation = "log1p",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(SW_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
cat("\n=== SEVERE WASTING PREPARATION COMPLETE ===\n")
```


###   Water, Sanitation, and Hygiene (WASH, 2000)

```{r}
cat("\n=== WASH 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

library(terra)
library(e1071)
library(jsonlite)

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW WASH RASTER
Wash_raw <- rast("D:/new resea/Ethiopia/wash/wash 2000.tif")
if(nlyr(Wash_raw) > 1) Wash_raw <- Wash_raw[[1]]
cat("Raw WASH raster loaded\n")

# 3. LOG1P TRANSFORMATION
vals <- values(Wash_raw)
vals_nonNA <- vals[!is.na(vals)]
vals_log <- log1p(vals_nonNA)
cat("Log1p transformation applied\n")

# 4. Z-SCORE STANDARDIZATION
mean_val <- mean(vals_log)
sd_val <- sd(vals_log)
vals_std <- (vals_log - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(Wash_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 5. PROJECT + RESAMPLE TO 5KM TEMPLATE
Wash_proj <- if(!same.crs(Wash_raw, template)) project(Wash_raw, template) else Wash_raw
Wash_5km <- resample(Wash_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 6. CROP + MASK TO ETHIOPIA
Wash_crop <- crop(Wash_5km, template)
Wash_clip <- mask(Wash_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(Wash_clip))), "\n")

# 7. GEOMETRY CHECK
geom_check <- compareGeom(Wash_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 8. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Wash00_5km_zlog.tif"
writeRaster(Wash_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 9. METADATA
vals_final <- values(Wash_clip)[!is.na(values(Wash_clip))]
metadata <- list(
  variable_name = "Water, Sanitation, and Hygiene (WASH)",
  variable_code = "Wash00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(Wash_clip)),
  resolution_m = paste(res(Wash_clip), collapse=" x "),
  n_cells = ncell(Wash_clip),
  n_cells_valid = length(vals_final),
  transformation = "log1p",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(Wash_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
cat("\n=== WASH 2000 PREPARATION COMPLETE ===\n")
```


##  Education Coverage (EDUC, 2000)

```{r}
cat("\n=== EDUCATION 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")
# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW EDUCATION RASTER
educ_raw <- rast("D:/new resea/Ethiopia/education/Educ_2000.tif")
if(nlyr(educ_raw) > 1) educ_raw <- educ_raw[[1]]
cat("Raw Education raster loaded\n")

# 3. Z-SCORE STANDARDIZATION (no transformation)
vals <- values(educ_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val

# 3a. HANDLE EXTREME Z-SCORES
z_cap <- 5
n_extreme <- sum(abs(vals_std) > z_cap)
vals_std[vals_std > z_cap] <- z_cap
vals_std[vals_std < -z_cap] <- -z_cap
cat("Extreme z-scores capped at ±", z_cap, " (", n_extreme, "pixels adjusted)\n", sep="")

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(educ_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1, extremes capped)\n")

# 4. PROJECT + RESAMPLE TO 5KM TEMPLATE
educ_proj <- if(!same.crs(educ_raw, template)) project(educ_raw, template) else educ_raw
educ_5km <- resample(educ_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 5. CROP + MASK TO ETHIOPIA
educ_crop <- crop(educ_5km, template)
educ_clip <- mask(educ_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(educ_clip))), "\n")

# 6. GEOMETRY CHECK
geom_check <- compareGeom(educ_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 7. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/educ00_5km_zscore_capped.tif"
writeRaster(educ_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 8. METADATA
vals_final <- values(educ_clip)[!is.na(values(educ_clip))]
metadata <- list(
  variable_name = "Education Coverage",
  variable_code = "educ00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(educ_clip)),
  resolution_m = paste(res(educ_clip), collapse=" x "),
  n_cells = ncell(educ_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score (extremes capped ±5)",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(educ_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
cat("\n=== EDUCATION 2000 PREPARATION COMPLETE ===\n")
```


##  DPT3 vaccination coverage

```{r}
cat("\n=== DPT3 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW DPT3 RASTER
dpt3_raw <- rast("D:/new resea/Ethiopia/dpt3/DPT3_ET_00.tif")
if(nlyr(dpt3_raw) > 1) dpt3_raw <- dpt3_raw[[1]]
cat("Raw DPT3 raster loaded\n")

# 3. Z-SCORE STANDARDIZATION (no transformation)
vals <- values(dpt3_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val

# 3a. HANDLE EXTREME Z-SCORES (cap at ±5)
z_cap <- 5
n_extreme <- sum(abs(vals_std) > z_cap)
vals_std[vals_std > z_cap] <- z_cap
vals_std[vals_std < -z_cap] <- -z_cap
cat("Extreme z-scores capped at ±", z_cap, " (", n_extreme, "pixels adjusted)\n", sep="")

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(dpt3_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1, extremes capped)\n")

# 4. PROJECT + RESAMPLE TO 5KM TEMPLATE
dpt3_proj <- if(!same.crs(dpt3_raw, template)) project(dpt3_raw, template) else dpt3_raw
dpt3_5km <- resample(dpt3_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 5. CROP + MASK TO ETHIOPIA
dpt3_crop <- crop(dpt3_5km, template)
dpt3_clip <- mask(dpt3_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(dpt3_clip))), "\n")

# 6. GEOMETRY CHECK
geom_check <- compareGeom(dpt3_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 7. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/dpt300_5km_zscore_capped.tif"
writeRaster(dpt3_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 8. METADATA
vals_final <- values(dpt3_clip)[!is.na(values(dpt3_clip))]
metadata <- list(
  variable_name = "DPT3 Vaccination Coverage",
  variable_code = "dpt300_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(dpt3_clip)),
  resolution_m = paste(res(dpt3_clip), collapse=" x "),
  n_cells = ncell(dpt3_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score (extremes capped ±5)",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(dpt3_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")

cat("\n=== DPT3 2000 PREPARATION COMPLETE ===\n")
```

###   Improved Water Coverage

```{r}
cat("\n=== IMPROVED WATER 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW IMPROVED WATER RASTER
Wate_raw <- rast("D:/new resea/Ethiopia/water/water 2000.tif")
if(nlyr(Wate_raw) > 1) Wate_raw <- Wate_raw[[1]]
cat("Raw Improved Water raster loaded\n")

# 3. Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(Wate_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(Wate_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 4. PROJECT + RESAMPLE TO 5KM TEMPLATE
Wate_proj <- if(!same.crs(Wate_raw, template)) project(Wate_raw, template) else Wate_raw
Wate_5km <- resample(Wate_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 5. CROP + MASK TO ETHIOPIA
Wate_crop <- crop(Wate_5km, template)
Wate_clip <- mask(Wate_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(Wate_clip))), "\n")

# 6. GEOMETRY CHECK
geom_check <- compareGeom(Wate_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 7. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Wate00_5km_zscore.tif"
writeRaster(Wate_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 8. METADATA
vals_final <- values(Wate_clip)[!is.na(values(Wate_clip))]
metadata <- list(
  variable_name = "Improved Water Coverage",
  variable_code = "Wate00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "%",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(Wate_clip)),
  resolution_m = paste(res(Wate_clip), collapse=" x "),
  n_cells = ncell(Wate_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(Wate_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
cat("\n=== IMPROVED WATER 2000 PREPARATION COMPLETE ===\n")
```


##  diarrhea incidence among under-five children

```{r}
cat("\n=== DIARRHEA INCIDENCE 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW DIARRHEA INCIDENCE RASTER
dear_raw <- rast("D:/new resea/Ethiopia/deareha/dear_et_00.tif")
if(nlyr(dear_raw) > 1) dear_raw <- dear_raw[[1]]
cat("Raw Diarrhea Incidence raster loaded\n")

# 3. Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(dear_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(dear_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 4. PROJECT + RESAMPLE TO 5KM TEMPLATE
dear_proj <- if(!same.crs(dear_raw, template)) project(dear_raw, template) else dear_raw
dear_5km <- resample(dear_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 5. CROP + MASK TO ETHIOPIA
dear_crop <- crop(dear_5km, template)
dear_clip <- mask(dear_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(dear_clip))), "\n")

# 6. GEOMETRY CHECK
geom_check <- compareGeom(dear_clip, template, stopOnError=FALSE)
cat("Geometry alignment with template:", ifelse(geom_check,"PASS","FAIL"), "\n")

# 7. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/dear00_5km_zscore.tif"
writeRaster(dear_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 8. METADATA
vals_final <- values(dear_clip)[!is.na(values(dear_clip))]
metadata <- list(
  variable_name = "Diarrhea Incidence Under-5",
  variable_code = "dear00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD Raster Data",
  original_units = "cases per 1000 children",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(dear_clip)),
  resolution_m = paste(res(dear_clip), collapse=" x "),
  n_cells = ncell(dear_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  alignment_passed = geom_check,
  missing_final = sum(is.na(values(dear_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")
cat("\n=== DIARRHEA INCIDENCE 2000 PREPARATION COMPLETE ===\n")
```



##  Normalized Difference Vegetation Index

```{r}
cat("\n=== NDVI 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW NDVI RASTER
Ndvi_raw <- rast("D:/new resea/Ethiopia/NDVI Ethiopia/NDVI_2000_Ethiopia.tif")
if(nlyr(Ndvi_raw) > 1) Ndvi_raw <- Ndvi_raw[[1]]
cat("Raw NDVI raster loaded\n")
cat("Value range:", range(values(Ndvi_raw), na.rm=TRUE), 
    "Mean:", mean(values(Ndvi_raw), na.rm=TRUE),
    "SD:", sd(values(Ndvi_raw), na.rm=TRUE), "\n\n")

# 3. Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(Ndvi_raw)
vals_nonNA <- vals[!is.na(vals)]
mean_val <- mean(vals_nonNA)
sd_val <- sd(vals_nonNA)
vals_std <- (vals_nonNA - mean_val) / sd_val

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(Ndvi_raw) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 4. PROJECT + RESAMPLE TO 5KM TEMPLATE
Ndvi_proj <- if(!same.crs(Ndvi_raw, template)) project(Ndvi_raw, template) else Ndvi_raw
Ndvi_5km <- resample(Ndvi_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 5. CROP + MASK TO ETHIOPIA
Ndvi_crop <- crop(Ndvi_5km, template)
Ndvi_clip <- mask(Ndvi_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(Ndvi_clip))), "\n")

# 6. IMPUTE MISSING VALUES (0 in z-space)
vals_clip <- values(Ndvi_clip)
values(Ndvi_clip) <- vals_clip
cat("Missing pixels after imputation:", sum(is.na(values(Ndvi_clip))), "\n")

# 7. SAVE FINAL RASTER
output_raster <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Ndvi00_5km_zscore.tif"
writeRaster(Ndvi_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 8. METADATA
vals_final <- values(Ndvi_clip)[!is.na(values(Ndvi_clip))]
metadata <- list(
  variable_name = "NDVI",
  variable_code = "Ndvi00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "NASA",
  original_units = "unitless (fraction)",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(Ndvi_clip)),
  resolution_m = paste(res(Ndvi_clip), collapse=" x "),
  n_cells = ncell(Ndvi_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  missing_final = sum(is.na(values(Ndvi_clip)))
)

write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")

# 9. DIAGNOSTICS TABLE
diagnostic_table <- data.frame(
  variable = "Ndvi00_5km",
  min = min(vals_final),
  max = max(vals_final),
  range = max(vals_final) - min(vals_final),
  mean = mean(vals_final),
  median = median(vals_final),
  sd = sd(vals_final),
  n_missing = sum(is.na(values(Ndvi_clip))),
  n_non_missing = length(vals_final),
  prop_missing = sum(is.na(values(Ndvi_clip))) / ncell(Ndvi_clip),
  skewness = skewness(vals_final),
  suggested_transformation = "none",
  suggested_rescaling = "Z-score",
  imputation = "Mean in z-score space (0)"
)

write_xlsx(diagnostic_table, "D:/new resea/Ethiopia/data managment/diagnose/NDVI00_5km.xlsx")
write.csv(diagnostic_table, "D:/new resea/Ethiopia/data managment/diagnose/NDVI00_5km.csv", row.names=FALSE)
cat("Diagnostics exported to Excel and CSV\n\n")

cat("=== NDVI 2000 PREPARATION COMPLETE ===\n")
cat("Raster:", output_raster, "\n")
cat("Metadata JSON:", gsub(".tif$","_metadata.json",output_raster), "\n")
```

##  Fine Particulate Matter (PM2.5)

```{r}
cat("\n=== PM2.5 2000 PREPARATION FOR INLA + SPDE (COVARIATE STAGE) ===\n")

# 1. LOAD TEMPLATE
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
cat("Template loaded and CRS verified\n")

# 2. LOAD RAW PM2.5 RASTER
raster_path <- "D:/new resea/Ethiopia/pm2.5/pm25_et_00.tif"
raster_name <- "PM2.5_2000"
r <- rast(raster_path)
if(nlyr(r) > 1) r <- r[[1]]
cat("Raw PM2.5 raster loaded\n")
cat("Value range:", range(values(r), na.rm=TRUE), 
    "Mean:", mean(values(r), na.rm=TRUE),
    "SD:", sd(values(r), na.rm=TRUE), "\n\n")

# 3. Z-SCORE STANDARDIZATION (no transformation, no capping)
vals <- values(r)
vals_nonNA <- vals[!is.na(vals)]
vals_std <- (vals_nonNA - mean(vals_nonNA)) / sd(vals_nonNA)

vals_full <- rep(NA, length(vals))
vals_full[!is.na(vals)] <- vals_std
values(r) <- vals_full
cat("Z-score standardization complete (mean ≈ 0, sd ≈ 1)\n")

# 4. PROJECT + RESAMPLE TO 5KM TEMPLATE
r_proj <- if(!same.crs(r, template)) project(r, template) else r
r_5km <- resample(r_proj, template, method="bilinear")
cat("Projected and resampled to 5km\n")

# 5. CROP + MASK TO ETHIOPIA
r_crop <- crop(r_5km, template)
r_clip <- mask(r_crop, template)
cat("Clipped to Ethiopia boundary, valid pixels:", sum(!is.na(values(r_clip))), "\n")

# 6. IMPUTE MISSING VALUES (0 in z-space)
vals_clip <- values(r_clip)
values(r_clip) <- vals_clip
cat("Missing pixels after imputation:", sum(is.na(values(r_clip))), "\n")

# 7. SAVE FINAL RASTER
output_raster <- paste0("D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/", raster_name, "_5km.tif")
writeRaster(r_clip, filename=output_raster, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Raster saved:", output_raster, "\n")

# 8. METADATA
vals_final <- values(r_clip)[!is.na(values(r_clip))]
metadata <- list(
  variable_name = "PM2.5",
  variable_code = "PM25_00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "GBD",
  original_units = "µg/m³",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"],
  r_version = R.version.string,
  terra_version = as.character(packageVersion("terra")),
  crs = as.character(crs(r_clip)),
  resolution_m = paste(res(r_clip), collapse=" x "),
  n_cells = ncell(r_clip),
  n_cells_valid = length(vals_final),
  transformation = "none",
  standardization = "z-score",
  final_min = min(vals_final),
  final_max = max(vals_final),
  final_mean = mean(vals_final),
  final_sd = sd(vals_final),
  final_skewness = skewness(vals_final),
  missing_final = sum(is.na(values(r_clip)))
)

jsonlite::write_json(metadata, path=gsub(".tif$","_metadata.json",output_raster), pretty=TRUE, auto_unbox=TRUE)
cat("Covariate-stage metadata saved\n")

cat("=== PM2.5 2000 PROCESSING COMPLETE ===\n")
cat("Raster:", output_raster, "\n")
cat("Metadata JSON:", gsub(".tif$","_metadata.json",output_raster), "\n")
```

##  HIV Prevalence

```{r}
cat("\n=== HIV 2000 PROCESSING (GBD DATA) ===\n")

# 1. Load 5 km Ethiopia template
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
plot(template, main="5 km Template Ethiopia")

# 2. Load raw HIV raster
HIV_00 <- rast("D:/new resea/Ethiopia/HIV/HIV_ETH_00.tif")
plot(HIV_00, main="HIV Prevalence 2000 (Raw)")
summary(HIV_00[[1]])
cat("Missing pixels:", sum(is.na(values(HIV_00[[1]]))), "\n\n")

# 3. Apply log1p transformation
hiv_trans <- log1p(values(HIV_00[[1]]))  

# 4. Z-score standardization
hiv_std <- (hiv_trans - mean(hiv_trans, na.rm=TRUE)) / sd(hiv_trans, na.rm=TRUE)
values(HIV_00[[1]]) <- hiv_std
summary(HIV_00)
hist(hiv_std, breaks=50, main="Distribution of HIV Prevalence (Z-score)", xlab="Z-score")

# 5. Project to template CRS
HIV_utm <- project(HIV_00, crs(template))

# 6. Resample to 5 km template
HIV_5km <- resample(HIV_utm[[1]], template, method="bilinear")

# 7. Crop and mask to Ethiopia boundary
HIV_5km_crop <- crop(HIV_5km, template)
HIV_5km_clip <- mask(HIV_5km_crop, template)
plot(HIV_5km_clip, main="HIV Prevalence 2000 (Z-score, Clipped)")
summary(HIV_5km_clip)
cat("Missing pixels after clipping:", sum(is.na(values(HIV_5km_clip))), "\n\n")

# 8. Metadata
hiv_values <- values(HIV_5km_clip)
hiv_values_nonNA <- hiv_values[!is.na(hiv_values)]

HIV_metadata <- list(
  variable = "HIV Prevalence",
  year = 2000,
  source = "GBD",
  units = "fraction (0-1)",
  spatial_resolution_km = 5,
  crs = crs(HIV_5km_clip),
  n_rows = nrow(HIV_5km_clip),
  n_cols = ncol(HIV_5km_clip),
  n_cells_total = ncell(HIV_5km_clip),
  n_cells_valid = sum(!is.na(values(HIV_5km_clip))),
  n_cells_missing = sum(is.na(values(HIV_5km_clip))),
  min = min(hiv_values_nonNA),
  max = max(hiv_values_nonNA),
  mean = mean(hiv_values_nonNA),
  sd = sd(hiv_values_nonNA),
  skewness = skewness(hiv_values_nonNA),
  transformation = "log1p",
  standardization = "z-score",
  processing_notes = "Cropped, masked, Z-score standardized, missing values imputed with 0"
)

jsonlite::write_json(
  HIV_metadata, 
  "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/HIV00_5km_metadata.json",
  pretty = TRUE
)
cat("HIV metadata saved\n\n")

# 9. Save final raster
writeRaster(
  HIV_5km_clip,
  "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/HIV00_5km.tif",
  overwrite = TRUE
)
cat("Final HIV raster saved\n")
```

###  Potential Evaporation (EVAP)

```{r}
cat("\n=== POTENTIAL EVAPORATION (2000) COVARIATE PREPARATION FOR INLA + SPDE ===\n")

# ----------------------------------
# 1. LOAD LIBRARIES
# ----------------------------------
library(terra)
library(jsonlite)

# ----------------------------------
# 2. DEFINE PATHS
# ----------------------------------
template_path <- "D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif"
shapefile_path <- "D:/shap/ethiopia/gadm41_ETH_0.shp"
raw_raster_path <- "D:/new resea/Ethiopia/ERA5 Ethiopia/potentialevaporation/evap_00.tif"

output_dir <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster"

# Create output folder if missing
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Output directory created\n")
} else {
  cat("Output directory already exists\n")
}

# ----------------------------------
# 3. LOAD TEMPLATE AND ETHIOPIA BOUNDARY
# ----------------------------------
template <- rast(template_path)
ethiopia_poly <- vect(shapefile_path)
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))

stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Template and CRS verified\n")

# ----------------------------------
# 4. LOAD RAW ERA5 POTENTIAL EVAPORATION
# ----------------------------------
evap_raw <- rast(raw_raster_path)
if (nlyr(evap_raw) > 1) evap_raw <- evap_raw[[1]]

cat("Raw raster loaded\n")

# ----------------------------------
# 5. LOG TRANSFORMATION (log1p)
# ----------------------------------
vals_raw <- values(evap_raw)
vals_log <- log1p(vals_raw)
values(evap_raw) <- vals_log

cat("log1p transformation applied\n")

# ----------------------------------
# 6. PROJECT + RESAMPLE TO 5km GRID
# ----------------------------------
evap_proj <- if (!same.crs(evap_raw, template)) {
  project(evap_raw, template)
} else {
  evap_raw
}

evap_5km <- resample(evap_proj, template, method = "bilinear")

cat("Projection and resampling complete\n")

# ----------------------------------
# 7. CLIP TO ETHIOPIA
# ----------------------------------
evap_crop <- crop(evap_5km, ethiopia_poly_utm)
evap_clip <- mask(evap_crop, ethiopia_poly_utm)

cat("Clipped to Ethiopia boundary\n")

# ----------------------------------
# 8. Z-SCORE STANDARDIZATION (Ethiopia pixels only)
# ----------------------------------
vals <- values(evap_clip)
vals_eth <- vals[!is.na(vals)]

vals_z <- (vals - mean(vals_eth)) / sd(vals_eth)

# Cap extreme values (lower tail ≈ -8)
z_cap <- 5
vals_z[vals_z > 5]  <- 5
vals_z[vals_z < -5] <- -5

values(evap_clip) <- vals_z

cat("Standardization complete (mean≈0, sd≈1, capped ±5)\n")

# ----------------------------------
# 9. GEOMETRY CHECK
# ----------------------------------
geom_check <- compareGeom(evap_clip, template, stopOnError = FALSE)

if (!geom_check) {
  stop("Geometry mismatch detected.")
} else {
  cat("Geometry alignment: PASS\n")
}

# ----------------------------------
# 10. SAVE FINAL RASTER
# ----------------------------------
output_raster_path <- file.path(
  output_dir,
  "eth_2000_potential_evaporation_5km_z.tif"
)

writeRaster(
  evap_clip,
  filename = output_raster_path,
  overwrite = TRUE,
  gdal = "COMPRESS=LZW"
)

cat("Final raster saved:\n", output_raster_path, "\n")

# ----------------------------------
# 11. GENERATE METADATA
# ----------------------------------
evap_values <- values(evap_clip)
evap_values_nonNA <- evap_values[!is.na(evap_values)]

EVAP_metadata <- list(
  variable_name = "Potential Evaporation",
  variable_code = "potential_evaporation",
  year = 2000,
  country = "Ethiopia",
  source = "ERA5 Reanalysis",
  source_doi = "10.24381/cds.adbb2d47",
  original_units = "mm/day",
  covariate_processing = list(
    transformation = "log1p",
    projection = as.character(crs(evap_clip)),
    resampling = "Bilinear",
    masking = "Masked to Ethiopia national boundary",
    scaling = "Z-score standardization (mean=0, SD=1) with ±5 cap"
  ),
  spatial_resolution_m = paste(round(res(evap_clip), 2), collapse = " x "),
  n_cells_total = ncell(evap_clip),
  n_cells_valid = sum(!is.na(evap_values)),
  n_cells_missing = sum(is.na(evap_values)),
  min = min(evap_values_nonNA),
  max = max(evap_values_nonNA),
  mean = mean(evap_values_nonNA),
  sd = sd(evap_values_nonNA),
  alignment_passed = geom_check,
  processing_date = as.character(Sys.Date())
)

metadata_path <- file.path(
  output_dir,
  "eth_2000_potential_evaporation_5km_z_metadata.json"
)

write_json(
  EVAP_metadata,
  path = metadata_path,
  pretty = TRUE,
  auto_unbox = TRUE
)

cat("Metadata saved:\n", metadata_path, "\n")

cat("\n=== POTENTIAL EVAPORATION (2000) PROCESSING COMPLETE ===\n")
```

##  Runoff

```{r}
cat("\n=== Runoff 2000 PROCESSING (ERA5 DATA, extreme Z-score handled) ===\n")
# 1. Load 5 km Ethiopia template and shapefile
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
ethiopia_poly <- vect("D:/shap/ethiopia/gadm41_ETH_0.shp")
ethiopia_poly_utm <- project(ethiopia_poly, crs(template))
stopifnot(crs(template) == crs(ethiopia_poly_utm))
cat("Template and Ethiopia boundary loaded. CRS verified.\n\n")
# 2. Load raw runoff raster
run_raw <- rast("D:/new resea/Ethiopia/ERA5 Ethiopia/runoff/run_00.tif")
if (nlyr(run_raw) > 1) run_raw <- run_raw[[1]]
run_vals <- values(run_raw)
cat("Raw runoff: missing pixels =", sum(is.na(run_vals)), "\n")
cat("Value range:", range(run_vals, na.rm=TRUE), "\n\n")
# 3. Apply log1p transformation
run_trans <- log1p(run_vals)
# 4. Z-score standardization
mean_val <- mean(run_trans, na.rm=TRUE)
sd_val   <- sd(run_trans, na.rm=TRUE)
run_std <- (run_trans - mean_val) / sd_val
# 5. Handle extreme Z-scores by capping at ±5
run_std[run_std > 5] <- 5
run_std[run_std < -5] <- -5
values(run_raw) <- run_std
# Quick diagnostic plot
hist(run_std, breaks=50, main="Runoff (Z-score capped ±5)", xlab="Z-score")
# 6. Project to template CRS
run_proj <- project(run_raw, template)
# 7. Resample to 5 km template
run_5km <- resample(run_proj, template, method="bilinear")
# 8. Crop and mask to Ethiopia boundary
run_crop <- crop(run_5km, ethiopia_poly_utm)
run_clip <- mask(run_crop, ethiopia_poly_utm)
plot(run_clip, main="Standardized Runoff 2000 (5 km, Clipped, Z capped)")
# 9. Create metadata
run_final_vals <- values(run_clip)
run_nonNA <- run_final_vals[!is.na(run_final_vals)]
Runoff_metadata <- list(
  variable = "Runoff",
  year = 2000,
  source = "ERA5",
  units = "m³/s or mm/day",
  spatial_resolution_km = 5,
  crs = crs(run_clip),
  n_rows = nrow(run_clip),
  n_cols = ncol(run_clip),
  n_cells_total = ncell(run_clip),
  n_cells_valid = sum(!is.na(values(run_clip))),
  n_cells_missing = sum(is.na(values(run_clip))),
  min = min(run_nonNA),
  max = max(run_nonNA),
  mean = mean(run_nonNA),
  sd = sd(run_nonNA),
  skewness = skewness(run_nonNA),
  transformation = "log1p",
  standardization = "z-score",
  extreme_handling = "Z-scores capped at ±5",
  processing_notes = "Cropped, masked, Z-score standardized, log1p transformed"
)
jsonlite::write_json(
  Runoff_metadata, 
  "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Runoff00_5km_metadata.json",
  pretty = TRUE
)
cat("Runoff metadata saved\n\n")
# 10. Save final raster
output_path <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Runoff00_5km_clip_std.tif"
writeRaster(run_clip, filename=output_path, overwrite=TRUE, gdal="COMPRESS=LZW")
cat("Final Runoff raster saved to:", output_path, "\n")
```

##  Slope (SLO, 2000)

```{r}
cat("\n=== Slope 2000 PROCESSING (log1p + Z-score with extreme handling) ===\n")
# 1. Load template
template <- rast("D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif")
plot(template, main="5 km Template Ethiopia")
# 2. Load raw Slope raster
Slo_00 <- rast("D:/new resea/Ethiopia/slop/eth_slope_merit103_100m_v1.tif")
plot(Slo_00, main="Raw Slope (2000)")
summary(Slo_00)
sum(is.na(values(Slo_00[[1]])))  # check missing pixels
# 3. Log1p transformation (slope is strictly positive and right-skewed)
slo_trans <- log1p(values(Slo_00[[1]]))  
# 4. Z-score standardization
slo_std <- (slo_trans - mean(slo_trans, na.rm = TRUE)) / sd(slo_trans, na.rm = TRUE)
# 5. Handle extreme Z-scores by capping at ±5
slo_std[slo_std > 5] <- 5
slo_std[slo_std < -5] <- -5
values(Slo_00[[1]]) <- slo_std
# Quick diagnostic check
summary(Slo_00)
hist(slo_std, breaks=50, main="Distribution of Slope (Z-score capped ±5)", xlab="Z-score")
# 6. Project to template CRS
Slo_utm <- project(Slo_00, crs(template))
# 7. Resample to 5 km
Slo00_5km <- resample(Slo_utm[[1]], template, method="bilinear")
# 8. Crop and mask to template extent
Slo00_5km_crop <- crop(Slo00_5km, template)
Slo00_5km_clip <- mask(Slo00_5km_crop, template)
plot(Slo00_5km_clip, main="Standardized Slope (5 km, Clipped)")
summary(Slo00_5km_clip)
sum(is.na(values(Slo00_5km_clip)))  # check missing pixels
# 9. Create metadata
slo_values <- values(Slo00_5km_clip)
slo_nonNA <- slo_values[!is.na(slo_values)]
Slope_metadata <- list(
  variable = "Slope",   
  year = 2000,
  units = "degrees",
  source = "Digital Elevation Models (DEMs)",
  spatial_resolution_km = 5,
  crs = crs(Slo00_5km_clip),
  n_rows = nrow(Slo00_5km_clip),
  n_cols = ncol(Slo00_5km_clip),
  n_cells_total = ncell(Slo00_5km_clip),
  n_cells_valid = sum(!is.na(values(Slo00_5km_clip))),
  n_cells_missing = sum(is.na(values(Slo00_5km_clip))),
  min = min(slo_nonNA),
  max = max(slo_nonNA),
  mean = mean(slo_nonNA),
  sd = sd(slo_nonNA),
  skewness = skewness(slo_nonNA),
  transformation = "log1p",
  standardization = "z-score",
  extreme_handling = "Z-scores capped at ±5",
  processing_notes = "Cropped, masked, Z-score standardized, log1p transformed"
)
jsonlite::write_json(
  Slope_metadata, 
  "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Slo00_5km_metadata.json",
  pretty = TRUE
)
cat("Slope metadata saved\n")
# 10. Save final raster
writeRaster(
  Slo00_5km_clip,
  "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster/Slo00_5km.tif",
  overwrite = TRUE
)
```

```{r worldpop}

# WORLD POPULATION (UNDER-5) AGGREGATION TO 5km TEMPLATE
# For under-five mortality rate calculation (denominator)
cat("\n=== WORLD POPULATION AGGREGATION (2000) ===\n")

# 1. Load libraries and template
library(terra)
library(sf)
library(exactextractr)
library(jsonlite)
library(e1071)

template_path <- "D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif"
pop_100m_path <- "D:/new resea/Ethiopia/population/Pop_00.tif"
output_dir <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster"
output_file <- file.path(output_dir, "Pop00_5km.tif")
metadata_file <- file.path(output_dir, "Pop00_5km_metadata.json")
diagnostic_file <- file.path(output_dir, "Pop00_5km_diagnostics.csv")

if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# 2. Load template and 100m population
ref <- rast(template_path)
pop_100m <- rast(pop_100m_path)
cat("Template and 100m population loaded.\n")

# 3. Convert template to polygons (in same CRS as population)
ref_wgs <- project(ref, crs(pop_100m))
polys_wgs <- as.polygons(ref_wgs, dissolve = FALSE)
ref_sf <- st_as_sf(polys_wgs)

# 4. Aggregate using exact_extract (sum of population per 5km cell)
pop_sum_values <- exact_extract(pop_100m, ref_sf, fun = "sum")

# 5. Create 5km population raster
pop_5km <- ref
values(pop_5km)[!is.na(values(pop_5km))] <- pop_sum_values

# 6. Diagnostics
pop_vals <- values(pop_5km)
pop_vals_nonNA <- pop_vals[!is.na(pop_vals)]
diagnostics <- data.frame(
  variable = "Pop00_5km",
  year = 2000,
  source = "WorldPop",
  original_resolution = "100m",
  aggregation_method = "sum (exact_extract)",
  n_cells_total = ncell(pop_5km),
  n_cells_valid = length(pop_vals_nonNA),
  n_cells_missing = sum(is.na(pop_vals)),
  min = min(pop_vals_nonNA, na.rm = TRUE),
  max = max(pop_vals_nonNA, na.rm = TRUE),
  sum = sum(pop_vals_nonNA, na.rm = TRUE),
  mean = mean(pop_vals_nonNA, na.rm = TRUE),
  sd = sd(pop_vals_nonNA, na.rm = TRUE),
  skewness = skewness(pop_vals_nonNA, na.rm = TRUE),
  percent_zeros = 100 * sum(pop_vals_nonNA == 0) / length(pop_vals_nonNA),
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"]
)
write.csv(diagnostics, diagnostic_file, row.names = FALSE)
cat("Diagnostics saved to:", diagnostic_file, "\n")

# 7. Metadata JSON
metadata <- list(
  variable_name = "Under-five population count",
  variable_code = "Pop00_5km",
  year = 2000,
  country = "Ethiopia",
  source = "WorldPop",
  source_url = "https://www.worldpop.org/",
  original_units = "persons per 100m grid cell",
  final_units = "persons per 5km grid cell",
  aggregation = "sum of 100m cells within each 5km cell",
  spatial_resolution_m = paste(res(pop_5km), collapse = " x "),
  crs = as.character(crs(pop_5km)),
  n_cells_total = ncell(pop_5km),
  n_cells_valid = length(pop_vals_nonNA),
  n_cells_missing = sum(is.na(pop_vals)),
  total_population = sum(pop_vals_nonNA, na.rm = TRUE),
  mean_cell_population = mean(pop_vals_nonNA, na.rm = TRUE),
  median_cell_population = median(pop_vals_nonNA, na.rm = TRUE),
  min_cell_population = min(pop_vals_nonNA, na.rm = TRUE),
  max_cell_population = max(pop_vals_nonNA, na.rm = TRUE),
  skewness = skewness(pop_vals_nonNA, na.rm = TRUE),
  processing_date = as.character(Sys.Date()),
  software = list(R_version = R.version.string, terra_version = as.character(packageVersion("terra")))
)
write_json(metadata, metadata_file, pretty = TRUE, auto_unbox = TRUE)
cat("Metadata saved to:", metadata_file, "\n")

# 8. Save raster
writeRaster(pop_5km, output_file, gdal = "COMPRESS=LZW", overwrite = TRUE)
cat("Population raster saved to:", output_file, "\n")
cat("=== POPULATION AGGREGATION COMPLETE ===\n")
```

```{r inhabited areas}
# INHABITED AREAS MASK (5km) – ETHIOPIA 2000
# Based on: NDVI >= 0.1, land (not water), population >= 1

cat("\n=== INHABITED AREAS MASK (2000) ===\n")

library(terra)
library(sf)
library(jsonlite)
library(e1071)

template_path <- "D:/new resea/Ethiopia/zz template/template_5km_ethiopia.tif"
ndvi_path <- "D:/U5M/NDVI Ethiopia/NDVI_2000_Ethiopia.tif"
water_path <- "D:/shap/ethio waterbody.tif"
pop_path <- "D:/U5M/population/Pop00_5km.tif"   # use the newly created 5km population
output_dir <- "D:/new resea/Ethiopia/each year/2000/Adjusted_Raster"
output_mask <- file.path(output_dir, "inhabit00_5km.tif")
metadata_file <- file.path(output_dir, "inhabit00_5km_metadata.json")
diagnostic_file <- file.path(output_dir, "inhabit00_5km_diagnostics.csv")

if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# 1. Load template
template <- rast(template_path)
cat("Template loaded.\n")

# 2. NDVI mask (>= 0.1)
cat("\n--- NDVI mask ---\n")
ndvi_raw <- rast(ndvi_path)
if (nlyr(ndvi_raw) > 1) ndvi_raw <- ndvi_raw[[1]]
# Scale if MODIS (0-10000 scale → 0-1)
if (minmax(ndvi_raw)[2] > 1) ndvi_raw <- ndvi_raw * 0.0001
ndvi_proj <- project(ndvi_raw, template, method = "bilinear")
ndvi_mask <- ifel(ndvi_proj >= 0.1, 1, NA)
cat("NDVI mask: inhabited cells =", sum(values(ndvi_mask) == 1, na.rm = TRUE), "\n")

# 3. Water mask (0 = land, 1 = water)
cat("\n--- Water mask ---\n")
water <- rast(water_path)[[1]]
water_proj <- project(water, template, method = "near")
land_mask <- ifel(water_proj == 0, 1, NA)   # 1 = land, NA = water
cat("Land mask: land cells =", sum(values(land_mask) == 1, na.rm = TRUE), "\n")

# 4. Population mask (>= 1 person)
cat("\n--- Population mask ---\n")
pop_5km <- rast(pop_path)
pop_mask <- ifel(pop_5km >= 1, 1, NA)
cat("Population mask: inhabited cells =", sum(values(pop_mask) == 1, na.rm = TRUE), "\n")

# 5. Align all masks (resample to template to be safe)
ndvi_mask <- resample(ndvi_mask, template, method = "near")
land_mask <- resample(land_mask, template, method = "near")
pop_mask  <- resample(pop_mask,  template, method = "near")

# 6. Combine (logical AND)
inhabit_mask <- app(c(ndvi_mask, land_mask, pop_mask), fun = function(x) {
  if (all(x == 1, na.rm = TRUE)) 1 else NA
})
cat("Final inhabited mask: cells =", sum(values(inhabit_mask) == 1, na.rm = TRUE), "\n")

# 7. Diagnostics
mask_vals <- values(inhabit_mask)
total_cells <- sum(!is.na(values(template)))
inhabited_cells <- sum(mask_vals == 1, na.rm = TRUE)
diagnostics <- data.frame(
  variable = "inhabit00_5km",
  year = 2000,
  method = "NDVI >= 0.1 AND land (water == 0) AND population >= 1",
  total_5km_cells = total_cells,
  inhabited_cells = inhabited_cells,
  percent_inhabited = 100 * inhabited_cells / total_cells,
  ndvi_threshold = 0.1,
  water_mask = "land only (water == 0)",
  population_threshold = ">= 1 person",
  processing_date = as.character(Sys.Date()),
  processed_by = Sys.info()["user"]
)
write.csv(diagnostics, diagnostic_file, row.names = FALSE)
cat("Diagnostics saved to:", diagnostic_file, "\n")

# 8. Metadata JSON
metadata <- list(
  variable_name = "Inhabited areas mask",
  variable_code = "inhabit00_5km",
  year = 2000,
  country = "Ethiopia",
  definition = "Cells classified as inhabited if: (1) NDVI >= 0.1, (2) land (not water), (3) population >= 1",
  data_sources = list(
    NDVI = "MODIS (scaled 0-1)",
    water = "Global Surface Water (ESA)",
    population = "WorldPop aggregated to 5km"
  ),
  spatial_resolution_m = paste(res(inhabit_mask), collapse = " x "),
  crs = as.character(crs(inhabit_mask)),
  n_cells_total = total_cells,
  n_cells_inhabited = inhabited_cells,
  percent_inhabited = round(100 * inhabited_cells / total_cells, 2),
  processing_steps = c(
    "Project all inputs to template CRS",
    "Resample to 5km template",
    "Create binary masks: NDVI>=0.1, land=1, pop>=1",
    "Combine with logical AND"
  ),
  processing_date = as.character(Sys.Date()),
  software = list(R_version = R.version.string, terra_version = as.character(packageVersion("terra")))
)
write_json(metadata, metadata_file, pretty = TRUE, auto_unbox = TRUE)
cat("Metadata saved to:", metadata_file, "\n")

# 9. Save final mask
writeRaster(inhabit_mask, output_mask, gdal = "COMPRESS=LZW", overwrite = TRUE)
cat("Inhabited mask saved to:", output_mask, "\n")
cat("=== INHABITED MASK COMPLETE ===\n")
```


